Treacle’s ability to form liquid phase condensates is essential for nucleolar fibrillar center assembly, efficient rRNA transcription and processing, and rRNA gene repair

Version 1

Abstract

We investigated the role of the nucleolar protein Treacle in organizing and regulating the nucleolus in human cells. Our results support Treacle’s capacity to form liquid-phase condensates through electrostatic interactions among molecules. The formation of these biomolecular condensates is crucial for segregating nucleolar fibrillar centers from the dense fibrillar component, as well as ensuring high levels of rRNA gene transcription and accurate pre-rRNA processing. The presence of both the central and C-terminal domains of Treacle is necessary for the creation of liquid-phase condensates. Initiation of phase separation is attributed to the C-terminal domain, while the central domain, characterized by repeated stretches of alternatively charged amino-acid residues, is vital for maintaining the condensate’s liquid state. When mutant forms of Treacle, incapable of forming liquid-phase condensates, are overexpressed in cells, it compromises the establishment of fibrillar centers, leading to the suppression of rRNA transcription and disruption of its processing. Additionally, these mutant forms fail to recruit TOPBP1, resulting in the suppression of the DNA damage response in the nucleolus.

Introduction

Applying polymer chemistry principles to biological molecules has significantly widened our understanding of membraneless compartments’ functioning (Mehta & Zhang, 2022; Mittag & Pappu, 2022). Multiple weak, cooperative, and dynamic interactions underlie the assembly of these self-organized structures, termed biomolecular condensates. Mechanistically, liquid–liquid phase separation (LLPS)—a process of de-mixing that yields a condensed protein-enriched phase and a dilute phase—can mediate the self-assembly of biomolecular condensates. In biological systems, phase separation is mainly a characteristic of multivalent or intrinsically disordered proteins (Banani et al, 2017; Shin & Brangwynne, 2017; Uversky, 2019). Biomolecular condensates serve as functional hubs in diverse cellular processes such as transcription (Sabari et al, 2020), microtubule nucleation (Woodruff et al, 2017), and adaptive stress response (Franzmann & Alberti, 2019).

Nucleolus, a large and complex subnuclear compartment, can also be considered a multicomponent and multilayered biomolecular condensate (Lafontaine et al, 2021). The nucleolus is formed around arrays of ribosomal gene repeats (rDNA), which are transcribed by RNA polymerase I (Pol I) to produce pre-ribosomal RNA (Cerqueira & Lemos, 2019). In the nucleolus, RNAs and hundreds of different proteins are segregated into three immiscible liquid phases: the fibrillar center (FC), which is the inner layer that contains the machinery for rDNA transcription by Pol I; the dense fibrillar component (DFC), which is the intermediate layer that is rich in fibrillarin; the granular component (GC), which is the outer layer that is rich in nucleophosmin (NPM1, B23). Although purified fibrillarin and NPM1 can separate into two distinct phases in vitro to create the droplet compartments of DFC and GC, respectively, the nature of the FC droplet compartment remains unclear (Feric et al, 2016). By examining the behavior of the FC in living cells, it has been proposed that the FC’s organization is a result of phase separation (Falahati et al, 2016; Falahati & Wieschaus, 2017). Furthermore, several studies have shown that the FC components relocate to the periphery of the nucleolus and aggregate to form large structures called nucleolar caps when Pol I transcription is inhibited by actinomycin D (AMD) or genotoxic drugs (Harding et al, 2015; Korsholm et al, 2019; Reynolds et al, 1964). This transformation process exhibits similarities to the fusion of liquid droplets. Treacle is the main molecule present in the FC, and it has both transcription-dependent and - independent functions (Gal et al, 2022). Its role in ribosome biogenesis is well-documented since it assists in the transcription and processing of rRNA (Gonzales et al, 2005; Lin & Yeh, 2009; Valdez et al, 2004). In addition, it is a master regulator of DNA damage response (DDR) in the nucleolus, mediating recruitment of repair factors to rDNA damage caused by IPpoI-induced breaks, replication stress, or R-loop formation (Ciccia et al, 2014; Korsholm et al, 2019; Larsen et al, 2014; Mooser et al, 2020; Velichko et al, 2021; Velichko et al, 2019). Thus, it acts as a nucleolar hub with both transcriptional and DNA repair functions. The reason for its multifunctionality can be attributed to its possible ability to undergo phase separation and engage in multivalent interactions. Protein structure predictors indicate that Treacle is one of the most intrinsically disordered proteins. In addition, Treacle’s partners in transcription and DNA repair (Pol I and TOPBP1, respectively) also utilize phase separation as a functioning mechanism (Frattini et al, 2021; Ide et al, 2020). Transcription of ribosomal genes can be regulated by phase separation of Pol I, while TOPBP1 condensation can activate the ATR signaling pathway (Frattini et al, 2021; Ide et al, 2020).

In this study, we present experimental evidence that Treacle can form biomolecular condensates, and we thoroughly characterize the structural determinants of Treacle regulating this process. Due to its ability for LLPS, Treacle facilitates the formation of nucleolar FCs, recruiting and concentrating transcription factors at rDNA, and separating FC components from the DFC. Collectively, this segregates rRNA synthesis and its subsequent processing in distinct compartments of the nucleolus. Disruption of Treacle’s LLPS capability, leading to the mixing of FC and DFC components, results in reduced efficiency of both processes, equivalent to the consequences of complete depletion of endogenous Treacle. Additionally, we demonstrate that Treacle’s phase separation is critical for its interaction with TOPBP1 and activation of DDR in rDNA during genotoxic stress. Our findings reveal the role of Treacle not only as a structural scaffold for FCs but also as a nucleolar hub integrating functions of ribosomal gene transcription, rRNA processing, and preserving rDNA integrity.

Treacle is a scaffold protein for nucleolar FCs and DFCs

First, we defined the role of Treacle in the structural organization of the nucleolus. We employed the CRISPR-Cas9 system to knock out the TCOF1 gene encoding Treacle in HeLa cells to obtain a population of cells lacking Treacle signals (Treacle-negative cells) (Fig. S1A). Immunocytochemical staining of UBF and RPA194 (FC markers), fibrillarin (DFC marker), and B23 and nucleolin (GC markers) revealed that Treacle-positive cells exhibited a classical tripartite nucleolar structure, with Treacle co-localizing with UBF1 and RPA194 (large sub-unit of Pol I) in FCs, fibrillarin in DFCs, and B23 and nucleolin surrounding FCs and DFCs (Fig. 1A). Treacle depletion led to the disappearance of distinct FC and DFC structures, causing RPA194, UBF, and fibrillarin to diffuse throughout the nucleolus and partially relocate to the nucleoplasm (Fig. 1B, S1A), whereas GC components remained unaffected (Fig. 1B). These results indicated that Treacle played a critical role in generating the integrated structure of FCs and DFCs.

Treacle is a scaffold protein for nucleolar FCs and DFCs.

(A) Intact HeLa cells (Treacle-positive) were fixed and co-immunostained with Treacle and with either RPA194, UBF, Fibrillarin, B23 or Nucleolin antibodies. DNA was stained with DAPI (gray). Cells were analyzed by laser scanning confocal microscopy. Representative images of cells and nucleoli (magnified images) are shown. Co-localization analysis was performed on the merged images. Graphs illustrate quantification in arbitrary units of fluorescence distribution along the lines shown in the figures. (B) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene. After 7-10 days after transfection, the cells were fixed and immunostained with Treacle and either RPA194, UBF, Fibrillarin, B23 or Nucleolin antibodies. Cells were analyzed by laser scanning confocal microscopy. Representative images of Treacle-negative cells and nucleoli (magnified images) are shown. (C) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene (experiment). After 7-10 days after transfection cell were fixed, immunostained with Treacle antibodies and subjected to cell sorting to obtain Treacle-negative populations. The sorted cell fractions were used for ChIP analysis with either Treacle, RPA194 or UBF antibodies. Intact HeLa cells were fixed, immunostained with IgG antibodies, passed through all FACS-related procedure and used as a control (Treacle-positive). ChIP was followed by qPCR using the d0, d1, d3, d5, d7 primers to the rRNA gene (positioned as indicated on the scheme). Data are represented relative to the input. Values are means ±SD from at least three independent replicates. (D) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene. After 7-10 days after transfection the cells were pulsed with EU (100 μM for 2 hr), fixed and immunostained with Treacle antibodies. EU (green) was revealed by click chemistry. The DNA was stained with DAPI (gray). Representative images of Treacle-positive and Treacle-negative HeLa cells are shown. Quantification of EU fluorescence intensities in Treacle-positive and Treacle-negative HeLa cells are show in the right panel (n>500). (E) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene. After 7-10 days after transfection the cells were fixed, immunostained with Treacle antibodies and subjected to cell sorting into Treacle-negative populations. The sorted cell fractions were used for RNA extraction. Intact HeLa cells were fixed, stained with IgG antibodies, passed through all FACS-related procedure and used as a control (Treacle-positive). RT-qPCR was performed; it shows levels of 47S pre-rRNA normalized to GAPDH mRNA. Normalized pre-rRNA level in Treacle-positive cells is set to 1. Values are mean ± SD. The calculation is presented for 5 biological replicates. (F) Hela cells were processed as described in (E). It shows levels of A’-site contained unprocessed rRNA normalized to GAPDH mRNA. Normalized unprocessed rRNA level in Treacle-positive cells is set to 1. Values are mean ± SD. The calculation is presented for 5 biological replicates.

Next, we investigated the impact of nucleolar FC disintegration on the efficiency of ribosomal gene transcription and the recruitment of Pol I transcription machinery to rDNA. Chromatin immunoprecipitation revealed reduced occupancy of UBF and RPA194 at the ribosomal gene coding region in Treacle-negative cells, indicating that Treacle amplifies the recruitment rather than initiates the primary binding of transcription machinery to rDNA (Fig. 1C). Depleting Treacle led to substantial suppression of ribosomal gene transcription, though not a complete shutdown, in contrast to the repression induced by AMD (Fig. 1D,E). Furthermore, cells with Treacle depletion also exhibit impairment in the processing of the 5′ external transcribed spacer, particularly in the cleavage of the A′ site (Fig. 1F) which is one of the initial steps in the 18S rRNA maturation.

In summary, Treacle functions as a scaffold protein in nucleolar FCs, and its depletion compromises both rRNA gene transcription and the processing of rRNA.

Treacle drives the formation of biomolecular condensates

Currently, it has become evident that various scaffold proteins can self-organize through the formation of liquid-phase condensates (Banani et al, 2017; Shin & Brangwynne, 2017). According to protein structure predictors (e.g., AlphaFold, IUPred2, PONDR, FuzDrop), Treacle is one of the most intrinsically disordered proteins (Fig. S2A). To investigate Treacle’s ability to self-organize into biomolecular condensates, we initially examined its LLPS properties in an in vitro system.

Purified recombinant Treacle underwent phase separation from the buffer and formed liquid droplets at low salt concentrations and in the presence of dextran as a crowding agent (Fig. 2A). Under these conditions, Treacle efficiently self-organized into large droplets, and their size positively correlated with the strength of crowding (Fig. 2A). These droplets exhibited a spherical shape and frequently fused with one another (Fig. 2B).

Treacle drives the formation of nuclear condensates.

(A) Purified recombinant Treacle underwent phase separation from the buffer and formed liquid droplets in the presence of 1-10% dextran as a crowding agent. The size of Treacle drops was measured (n>500). (B) Drops of recombinant Treacle were time-lapse imaged. Representative images of drop fusion events are shown. (C) HeLa cells were transfected with Treacle-FusionRed-Cry2 (opto-Treacle) construct. Formation of opto-Treacle drops was induced with blue light illumination of the cells for 10 sec. DNA was stained with Hoechst 33342 (gray). Representative images are shown. (D) HeLa cells were transfected with Treacle-Katushka2S (Treacle-2S) at a quantity of 50 ng plasmids per 2×105 cells. For low or high levels of expression analysis, the cells were fixed 16-24 or 48 h after transfection, respectively. DNA was stained with DAPI (gray). Cells were analyzed by laser scanning confocal microscopy. Surface reconstructions of Treacle-2S drops are shown in the right panel. Graphs illustrate quantification in arbitrary units of the 3D analysis of Treacle-2S drops’ shape by LasX software. Sphere form factor = sphere area/particle area. (E) 16-24 h after transfection with Treacle-2S, HeLa cells were treated with 0.05µg/ml actinomycin D (AMD) to induce rDNA transcriptional repression and time-lapse imaged for 120 minutes. (F) 48 h after transfection with Treacle-2S, HeLa cells were live time-lapse imaged. Representative images of cells with Treacle-2S drops fusion events are shown. Graphs illustrate the frequency of fusion Treacle condensates per cell for 12 hours. The calculation is presented for 6 biological replicates of 10 cells each. (G) HeLa cells were transiently transfected with Treacle-2S. 16-24 h after transfection FRAP analysis of Treacle-labelled fibrillar center (FCs) dynamics was performed. Part of one FC was bleached, and fluorescence recovery was monitored. Representative time-lapse images of the photobleached FC are shown (magnified images). (H) 16-24 h after transfection with Treacle-2S, HeLa cells were treated with 0.05µg/ml AMD to induce the formation of nucleolar caps. Part of the nucleolar cap was bleached, and fluorescence recovery was monitored. Representative time-lapse images of photobleached nucleolar caps are shown (magnified images). (I) Graphs illustrate the quantification of the Treacle-2S FRAP analysis described in (G) and (H). Each trace represents an average of measurements for at least twenty FCs or nucleolar caps; error bars represent SD. (J) HeLa cells were transiently transfected with Treacle-2S. 48 h after transfection, FRAP analysis of Treacle-2S extranucleolar condensates dynamics was performed. Part of one drop was bleached, and fluorescence recovery was monitored. Representative time-lapse images of photobleached Treacle-2S extranucleolar condensate are shown (magnified images). (K) HeLa cells were transiently transfected with Treacle-2S. 72-96 h after transfection, HeLa cells were processed as described in (J). (L) Graphs illustrate the quantification of the Treacle-2S FRAP analysis described in (J) and (K). Each trace represents an average of measurements for at least twenty Treacle-2S condensates; error bars represent SD.

To investigate Treacle’s ability to self-organize into biomolecular condensates within cells, we generated fusion proteins by linking Treacle to FusionRed (a non-self-dimerizing fluorescent protein) and Arabidopsis thaliana cryptochrome 2 (Cry2), known to oligomerize upon exposure to blue light, facilitating the formation of "optoDroplets" (Shin & Brangwynne, 2017). Expression of this opto–Treacle chimeric protein in HeLa cells revealed foci formation in nucleoli in the absence of blue light, consistent with Treacle’s natural tendency to occupy FCs (Fig. 2C; controls in Fig. S2B). Upon blue light exposure, opto– Treacle formed multiple small foci throughout the nucleoplasm (Fig. 2C). We next explored Treacle’s ability to form biomolecular condensates when overexpressed as a fusion with the fluorescent protein Katushka2S (referred to as 2S) (Shcherbo et al, 2007). Initially, at low expression levels (16-24 hours post-transfection), Treacle’s fusion protein formed foci only in nucleoli reflecting its natural occupation of FCs (Fig. 2D top panel). However, with increased expression levels (48-72 hours post-transfection), it began to form large spherical structures in the nucleoplasm (Fig.2D). Both intranucleolar and extranucleolar Treacle–2S foci, induced by low and high levels of fusion protein expression respectively, exhibited a spherical shape with an aspect ratio close to one, suggesting susceptibility to surface tension (Fig. 2D). Transmission electron microscopy confirmed the internal homogeneity of Treacle droplets, showing circular structures with homogeneous protein content without any electron-dense or -light regions inside the droplets (Fig. S2C).

Similar to the in vitro-formed Treacle condensates, intracellular Treacle condensates demonstrated the ability to fuse. Our observations reveal that during AMD-induced (Fig. 2E, Movie S1) or DNA damage-induced (Fig. S2D, Movie S2) rDNA transcriptional repression, nucleolar caps form due to the fusion of intranucleolar Treacle–2S foci. Prolonged live-cell imaging further demonstrated that extranucleolar biomolecular condensates from highly overexpressed Treacle–2S also frequently fuse, indicating dynamic clustering (Fig. 2F, Movie S3).

Liquid condensates exhibit a high molecular exchange rate, often assessed through fluorescence recovery after photobleaching (FRAP) (Ganser & Myong, 2020). FRAP analysis of the entire FC showed slow fluorescence recovery of Treacle, likely due to limited molecule diffusion between different FCs (Fig. S2E). However, when bleaching a specific region within the FC, rapid fluorescence recovery was observed, indicating liquid-like dynamics (Fig. 2G, I). We then analyzed the molecular exchange rate of Treacle in nucleolar caps induced by AMD (Fig. 2H, I) and in extranucleolar droplets induced by high-level expression of Treacle-2S (Fig. 2J, L). Both types of condensates exhibited dynamics consistent with liquid behavior, with slightly higher rates than in FCs. Fluorescence recovery in FCs, nucleolar caps, and extranucleolar condensates never reached the initial values over the analyzed time periods. This suggests that the high molecular exchange rate occurs through the mixing of Treacle molecules within the condensate boundaries and does not involve external diffusion. It is important to note that liquid dynamics were characteristic of Treacle droplets formed 16-48 hours post-transfection, while FRAP analysis of protein dynamics in large extranucleolar droplets 72-96 hours after transfection revealed a shift towards gel-like behavior of the observed condensates (Fig. 2 K,L). Presumably, Treacle condensates undergo a liquid-to-gel phase transition over time or upon reaching a critical protein concentration within the condensate. Thus, using diverse model systems, we illustrated that Treacle can indeed create biomolecular condensates exhibiting the primary features of structures mediated by LLPS.

Treacle’s phase separation is regulated by its central and C-terminal domains

We next searched the structural determinants of Treacle contributing to its LLPS characteristics. Despite being a low-complexity protein, Treacle consists of three functional domains: the N-terminal domain (ND) containing the LisH motif, the C-terminal domain (CD) which includes nuclear (NLS) and nucleolar (NOLC) localization signals, and the central or Repeated Domain (RD) consisting of 14 low complexity regions, 11 of which are homologous to each other (Fig. 3A). To identify the domains contributing to LLPS, we generated Treacle mutants with deleted ND (Δ1-83), RD (Δ83-1121), or CD (Δ1121-1488), and overexpressed them in HeLa cells. We decided to overexpress the deletion mutants in wild-type cells and assess their LLPS capacity using several intracellular condensate models: (i) formation of FCs and (ii) nucleolar caps under low expression levels (24 hours post-transfection), and (iii) formation of large extranucleolar condensates under high expression levels (48 hours post-transfection).

Treacle’s phase separation is regulated by its central and C-terminal domains

(A) The structure of the most common isoform of Treacle. Treacle isoform d (1488 amino acids, 152 kDa, NP_001128715.1) is encoded by the Treacle transcript variant 4. It is an intrinsically disordered protein with N-terminal (ND, 1–83 aa), C-terminal (CD, 1121–1488 aa) regions, and central repeated domain (RD, 83–1121 aa) consisting of 14 low complexity regions (LCR). (B) HeLa cells were transfected with 2S-fused full-length Treacle. For low or high levels expression analysis cells were cultivated 16-24 or 48 h after transfection respectively. The expression level is indicated by the colored zone to the left of the cell images. HeLa cells with low expression level were additionally treated with CX-5461 to induce rDNA transcriptional repression DNA and subsequent nucleolar cap formation. Cells were fixed and analyzed by laser scanning confocal microscopy. DNA was stained with DAPI (gray). Representative images of full-length Treacle-2S condensate are shown. (C) HeLa cells were transiently transfected 2S-fused full-length Treacle and cultivated as described in (B). FRAP analysis of 2S-fused full-length Treacle condensates dynamics was performed. Graphs illustrate quantification of FRAP analysis for each type of Treacle-formed condensates (FCs, nucleolar caps, extranucleolar condensates). Each trace represents an average of measurements for at least twenty for each type of condensates; error bars represent SD. (D) HeLa cells were transfected with 2S-fused Treacle Δ1-83 deletion mutant and processed as described in (B). (E) HeLa cells were transfected with 2S-fused Treacle Δ1-83 deletion mutant and cultivated as described in (B). FRAP analysis of 2S-fused Treacle Δ1-83 condensates dynamics was performed as described in (C). (F) HeLa cells were transfected with 2S-fused Treacle Δ83-1121 deletion mutant and processed as described in (B). (G) HeLa cells were transfected with 2S-fused Treacle Δ83-1121 deletion mutant and cultivated as described in (B). FRAP analysis of 2S-fused Treacle Δ83-1121 condensates dynamics was performed as described in (C). (H) HeLa cells were transfected with Treacle Δ1121-1488 deletion mutant fused with 2S and nuclear localization signal (NLS) from simian virus 40 (SV40). For low or high levels expression analysis cells were cultivated 16-24 or 48 h after transfection respectively. The expression level is indicated by the colored zone to the left of the cell images. Cells were fixed and analyzed by laser scanning confocal microscopy. DNA was stained with DAPI (gray). Representative images of cells are shown. (I) Purified recombinant Treacle underwent phase separation in the presence of 10% 1,6-hexandiol (1,6-HEX) or 500 mM sodium chloride. In each case, 10% dextran was used as a crowding agent. The size of Treacle drops was measured (n>500). (J) HeLa cells were transfected with 2S-fused full-length Treacle. For low or high levels expression analysis, cells were cultivated 16-24 or 48 h after transfection respectively. The expression level is indicated by the colored zone to the left of the cell images. Cells were treated with 5% 1,6-hexandiol (1,6-HEX) for 10 min or 200 mM ammonium acetate for 5 min. Cells were fixed and analyzed by laser scanning confocal microscopy. DNA was stained with DAPI (gray). Representative images of cells are shown. (K) Charge plots of full-length Treacle (top panel) and charge-scrambled Treacle (Treacle CS) form (bottom panel) are shown. Positive and negative charge blocks are depicted by blue and red triangles, respectively. Charge distribution was calculated as the sum of the charges (Arg and Lys, +1; Glu and Asp, −1;) in the 25 amino acids window range. The center of the panel shows the aligned amino acid sequences of the third LCR of RD for the full Treacle and Treacle CS. (L) HeLa cells were transfected with charge-scrambled 2S-fused Treacle (Treacle-2S CS) and processed as described in (B). (M) HeLa cells were transfected with Treacle-2S CS and cultivated as described in (B). FRAP analysis of Treacle-2S CS condensates dynamics was performed as described in (C). (N) HeLa cells were transfected with 2S-fused Treacle Δ1350-1488 deletion mutant (Treacle-2S ΔNOLC) and processed as described in (B). (O) HeLa cells were transfected with Treacle-2S ΔNOLC and cultivated as described in (B). FRAP analysis of Treacle-2S ΔNOLC condensates dynamics was performed as described in (C).

The Δ1-83 mutant demonstrated LLPS properties indistinguishable from full-length Treacle: it concentrated in FCs, formed stable nucleolar caps during transcriptional repression, and, upon increased expression levels, generated numerous large extranucleolar condensates (Fig. 3B, D). Similar to full-length Treacle, Δ1-83 exhibited a high rate of molecular exchange in all types of model condensates (Fig. 3C, E, S3A-C).

The mutant lacking the central domain (Δ83-1121) exhibited LLPS properties distinct from full-length Treacle. Under low expression levels, it failed to concentrate in FCs (Fig. 3F) and diffusely localized in the nucleoplasm (Fig. 3F) while still maintaining a high molecular dynamics rate (Fig. 3G, S3D). During rDNA transcriptional repression, Δ83-1121 did not form classical nucleolar caps but partially redistributed between the nucleolar periphery and nucleoplasm, where it began to form condensates (Fig. 3F). Surprisingly, such relocation was associated with a change in its phase state. FRAP analysis demonstrated that both perinucleolar and nucleoplasmic condensates of Δ83-1121, formed due to nucleolar transcriptional repression, began to transition into a solid state (liquid-solid phase separation,LSPS) (Fig. 3G, S3E-F). It is possible, that without the central domain, Treacle can maintain LLPS only through heterotypic interactions with rRNA. A similar liquid-solid transition was observed for Δ83-1121 under high expression levels. In this case, it no longer formed distinct extranucleolar condensates but merged into a unified pan-nuclear solid-state network (Fig. 3F, G, S3G).

The Δ1121-1488 mutant could not translocate into the nucleus due to the absence of its own NLS (data not shown), so it was additionally fused with the NLS from the SV40 virus. Interestingly, in this form, Δ1121-1488 began to localize in the nucleus but failed to form condensates even under high expression levels (Fig. 3H).

These results suggest that it is the RD that provides the liquid behavior of the Treacle. However, RD is not capable of autonomous condensation and utilizes the CD as a platform for its own nucleation. In turn, CD cannot phase separate independently and requires either RD or rRNA, without which it undergoes a LSPS. Next, we investigated the features of the amino acid sequences of RD and CD that could contribute to these properties.

We started by analyzing the type of low-complexity region (LCR) interactions in full-length Treacle condensates. We first examined the assembly of Treacle droplets in vitro in the presence of 1,6-hexanediol, high salt concentrations, or at different pH values. The efficiency of Treacle droplet formation in vitro remained unchanged in the presence of 1,6-hexanediol (Fig. 3I), whereas increasing salt concentration blocked droplet formation (Fig. 3I). Additionally, in vitro, condensation of Treacle was sensitive to changes in pH. The droplet size significantly decreased under acidic conditions (pH 5.5) or alkalization of the environment (pH 10), and was maximal at pH 8.5, corresponding to the isoelectric point for Treacle (Fig. S3H). Intracellular Treacle condensates were also sensitive to changes in ionic strength. Treatment with the cell-permeable salt ammonium acetate disrupted both FCs and extranucleolar condensates formed at low and high Treacle expression levels, respectively (Fig. 3J). In contrast, treatment with 1,6-hexanediol did not have a similar effect (Fig. 3J). Thus, it is evident that Treacle’s LLPS is driven by electrostatic, but not hydrophobic, interactions Analysis of the charge distribution along the amino acid sequence of Treacle revealed that the repeats in the central domain constitute strong diblock ampholytes, where a block of positive charge is followed by a block of negative charge (Fig. 3K). To determine the physiological importance of this charge distribution, a variant of Treacle, Treacle CS, was created with the same overall net charge but with scrambled blocks (Fig. 3K). In Treacle CS, regions of opposite charge were removed while maintaining the same overall isoelectric point, amino acid composition, and positions of all other residues. As anticipated, scrambling the charges in RD completely blocked Treacle’s phase separation ability. Similar to Δ83-1121, CS was unable to form FCs or nucleolar caps (Fig. 3L), remained in a liquid state only in the presence of rRNA (Fig. 3M), and underwent a liquid-solid phase transition during nucleolar transcriptional repression or high levels of overexpression (Fig. 3L, M, S3I-K). This indicates the importance of the charge distribution pattern in RD for maintaining the phase separation properties of Treacle.

The C-terminal domain of Treacle exhibited a significantly positive charge, attributed to the K-rich region at the C-terminus, which is also a NOLC. To analyze the significance of this charge, a Treacle variant with a deletion of the 1350-1488 region (ΔNOLC), including most lysines in CD, was created. As expected, this form of Treacle did not localize to the nucleolus but, nonetheless, was capable of forming individual perinucleolar condensates (Fig. 3N). These condensates demonstrated a liquid state, and the molecular exchange rate within them was even slightly higher than in condensates of full-length Treacle (Fig. 3O, S3L-N). Interestingly, with an increase in the expression level of ΔNOLC, a significant increase in the volume of such condensates was observed, but not an increase in their number (Fig. 3N). This suggests that positively charged lysines in CD serve as nucleation points for autonomous Treacle condensate formation.

In summary, the results indicate that the central and C-terminal domains cooperatively control the phase separation properties of Treacle, while the N-terminal domain does not make a substantial contribution. The highly positively charged CD serves as the nucleation center for RD but exhibits ambivalent phase properties, transitioning from LLPS to LSPS in the absence of rRNA. In turn, RD, with alternating charge blocks in the repeating domains, maintains Treacle condensates in a stably liquid state, thus counteracting rRNA-dependent phase fluctuations in CD. The combination of CD and RD supports a balance of viscoelastic properties in Treacle condensates.

Treacle LLPS is essential for the transcription and processing of rRNA

In the previous part of the study, we demonstrated that Treacle supports the transcription and processing of rRNA. We hypothesized that this functionality of Treacle could be attributed to its ability for LLPS. To test this hypothesis, we significantly downregulated the endogenous Treacle expression using RNA interference (Fig. S4A) and then analyzed the structure of FC/DFC and rRNA transcription levels in the context of the expression of full-length or LLPS-defective (Δ83-1121 or CS) forms of Treacle. It was observed that compared to the full-length Treacle, which formed classical FC/DFC, each of the LLPS-defective mutants induced dispersion and mixing of FC and DFC components (Fig.4 A,C), similar to what occurs during CRISPR/Cas9-mediated depletion of endogenous Treacle (Fig. 4B,D). In this process, GC did not undergo visible changes (Fig. S4B-D). Additionally, we found that the expression of LLPS-defective mutants significantly reduced the transcription level of rRNA compared to full-length Treacle (Fig.4E). This effect could not be explained by a possible loss of the ability of LLPS-defective mutants of Treacle to bind to rDNA, as both LLPS-defective mutants occupied the coding region of ribosomal repeats in a manner similar to full-length Treacle (Fig. 4F).

Treacle LLPS is essential for the transcription and processing of rRNA

(A) Endogenous Treacle was depleted by siRNA-mediated knockdown (Treacle kd). Next, Treacle-depleted cells were transfected with siRNA-resistant 2S-fused full-length Treacle (Treacle-2S full), charge-scrambled Treacle (Treacle-2S CS) or Treacle Δ83-1121 deletion mutant (Treacle-2S Δ83-1121). Cells were fixed 16-24 after transfection and immunostained with anti-UBF antibodies. Cells were analyzed by laser scanning confocal microscopy. Representative images of magnified nucleoli are shown. Co-localization analysis was performed on the merged images. Graphs illustrate quantification in arbitrary units of Treacle-2S variants and UBF fluorescence distribution along the lines shown in the figures. (B) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene. After 7-10 days after transfection, the cells were fixed and immunostained with Treacle and UBF antibodies. Cells were analyzed by laser scanning confocal microscopy. Representative images of Treacle-negative nucleolus (magnified images) are shown. (C) HeLa cells were processed as described in (A) and immunostained with Fibrillarin antibodies. (D) HeLa cells were processed as described in (B) and immunostained with Fibrillarin and UBF antibodies. Cells were analyzed by laser scanning confocal microscopy. Representative images of Treacle-negative nucleolus (magnified images) are shown. (E) HeLa cells were processed as described in (A). Cells were fixed 16-24 after transfection and subjected to cell sorting into 2S-positive populations. The sorted cell fractions were used for RNA extraction. RT-qPCR was performed; it shows levels of 47S pre-rRNA normalized to GAPDH mRNA. Normalized pre-rRNA level in 2S-fused full-length Treacle-positive cells is set to 1. Values are mean ± SD. The calculation is presented for 5 biological replicates. (F) HeLa cells were processed as described in (A). Cells were fixed 16-24 after transfection and subjected to ChIP-seq analysis with Katushka2S antibodies. ChIP-seq signal were normalized to the input. (G) HeLa cells were processed as described in (A). Cells were fixed 16-24 after transfection and stained for rRNA (revealed by single-molecule FISH, smFISH). Cells were analyzed by laser scanning confocal microscopy. Representative images of magnified nucleoli are shown. Co-localization analysis was performed on the merged images. Graphs illustrate quantification in arbitrary units of Treacle-2S variants and smFISH fluorescence distribution along the lines shown in the figures. (H) HeLa cells were processed as described in (B) and stained with UBF antibodies and smFISH. Representative images of Treacle-negative nucleolus (magnified images) are shown. (I) HeLa cells were processed as described in (A). Cells were fixed 16-24 after transfection and subjected to cell sorting into 2S-positive populations. The sorted cell fractions were used for RNA extraction. RT-qPCR was performed; it shows levels of A’-site contained unprocessed rRNA normalized to GAPDH mRNA. Normalized unprocessed rRNA level in 2S-fused full-length Treacle-positive cells is set to 1. Values are mean ± SD. The calculation is presented for 5 biological replicates.

Thus, this observation suggests that the LLPS properties of Treacle provide its ability to concentrate components of the transcriptional machinery in FC and separate them from DFC. Moreover, such separation is likely required to maintain highly efficient transcription of ribosomal genes. We hypothesized that the same principle could regulate the role of Treacle in the processing of rRNA. To test this assumption, we depleted the level of endogenous Treacle using RNA interference (Fig. S4A) and then analyzed the spatial localization of newly synthesized 47S-rRNA and its processing level in cells expressing miRNA-resistant full-length or LLPS-defective (Δ83-1121 or CS) forms of Treacle. Microscopy analyses with FISH-labeled rRNA revealed that newly synthesized 47S-rRNA transcript clustered at the periphery of FCs in full-length Treacle-expressing cells, reflecting the radial flow of rRNA from FC to GC (Fig. 4G). However, in cells expressing LLPS-defective Treacle mutants, no clear clustering of newly synthesized rRNA was observed. Instead, the newly synthesized transcripts were diffusely mixed with Δ83-1121 and CS Treacle (Fig. 4G). In cells with depleted endogenous Treacle, nascent pre-rRNA transcripts were similarly mixed with delocalized UBF (Fig. 4H). Finally, reverse transcription-qPCR analysis confirmed that cells overexpressing LLPS-defective Treacle mutants demonstrated impairment of 5′ external transcribed spacer processing compared to cells overexpressing full-length Treacle (Fig.4I). Thus, it can be concluded that the mixing of FC and DFC components induced by the expression of LLPS-defective Treacle, as well as the depletion of endogenous Treacle, disrupts the directional traffic of nascent pre-rRNA from FC to DFC and further to GC, which may be a cause of inefficient processing.

Our results indicate that Treacle’s phase separation properties provide separation of FC and DFC, leading to the spatial segregation of rRNA synthesis and its subsequent processing. The mixing of FC and DFC components due to the disruption of Treacle’s LLPS ability results in reduced efficiency of both processes, equivalent in consequences to the complete depletion of endogenous Treacle.

Treacle phase separation is essential for DDR activation in ribosomal genes under genotoxic stress conditions

Previous studies have demonstrated that Treacle is critically required for the induction of DDR in ribosomal genes under certain types of stress (Korsholm et al, 2019; Larsen et al, 2014; Mooser et al, 2020; Velichko et al, 2021; Velichko et al, 2019). Here, we aimed to investigate the contribution of Treacle’s phase separation to rDNA damage response. To induce rDNA damage, we used the widely used chemotherapeutic drug etoposide (VP16), which acts as a topoisomerase II poison inducing double-strand breaks (DSBs) (Bax et al, 2019). Treatment of cells with 90 µM VP16 for 30 minutes resulted in the rapid recruitment of TOPBP1 to nucleoli and its colocalization with Treacle (Fig. 5A). Furthermore, proximity ligation assay (PLA) analysis conducted after DNA damage induction provided evidence of a physical association between TOPBP1 and Treacle (Fig. 5B). The reduction of endogenous Treacle levels, achieved either through RNA interference (Treacle kd) or CRISPR/Cas9-mediated depletion (Treacle kn), effectively blocked the relocalization of TOPBP1 to nucleoli and its occupancy of rDNA (Fig. 5C, S5A). Interestingly, with RNA interference, a substantial amount of Treacle still remained in nucleoli (Fig. S5A). However, this amount was insufficient to facilitate its interaction with TOPBP1, likely due to the need for specific stoichiometric ratios for efficient interaction.

Treacle phase separation is essential for DDR activation in ribosomal genes under genotoxic stress conditions

(A) Control (DMSO-treated) and VP16-treated (90 μM, 30 min) HeLa cells were stained for Treacle (magenta) and TOPBP1 (green) and analyzed by laser scanning confocal microscopy. The DNA was stained with DAPI (gray). Co-localization analysis was performed on the merged images (magnified images). Graphs illustrate quantification in arbitrary units of Treacle and TOPBP1 fluorescence distribution along the lines shown in the figures. (B) Control (DMSO-treated) and VP16-treated (90 μM, 30 min) HeLa cells were subjected to PLA with antibodies against TOPBP1 and Treacle. DNA was stained with DAPI (gray). PLA detection of Treacle-TOPBP1 interactions is visible as distinct green, fluorescent dots. (C) Сontrol (mock-treated) HeLa cells or cells with siRNA-mediated Treacle knockdown (Treacle kd) were treated with DMSO or VP16 (90 μM, 30 min). ChIP experiments were performed with antibodies against TOPBP1. ChIP was followed by qPCR using the d1 primers to the promoter of the rRNA gene (positioned as indicated on the scheme). Data are represented relative to the input. Values are means ±SD from at least three independent replicates. (D) Endogenous Treacle was depleted by siRNA-mediated knockdown of the TCOF1 gene (Treacle kd). Next, Treacle-depleted cells were transfected with plasmid constructs encoding siRNA-resistant 2S-fused full-length Treacle, 2S-fused Treacle Δ83-1121, or 2S-fused charge-scrambled mutant Treacle (Treacle-2S CS). 24 h after transfection, cells were treated with VP16 (90 μM for 30 min), fixed and stained for TOPBP1 (green) and analyzed by laser scanning confocal microscopy. The DNA was stained with DAPI (gray). (E) Endogenous Treacle was depleted by siRNA-mediated knockdown of the TCOF1 gene (Treacle kd). Treacle-depleted cells were transfected with plasmid constructs encoding siRNA-resistant 2S-fused full-length Treacle, 2S-fused Treacle Δ83-1121, or 2S-fused charge-scrambled mutant Treacle (Treacle-2S CS). 24 h after transfection, cells were treated with DMSO or VP16 (90 μM for 30 min) and fixed. Cells were subjected to cell sorting into 2S-positive populations. At least 2×106 sorted cells were used for ChIP with TOPBP1 antibodies. ChIP was followed by qPCR using the d1 primers to the promoter of the rRNA gene (positioned as indicated on the scheme). Data are represented relative to the input. Values are means ±SD from at least three independent replicates. (F) Intact HeLa cells and cells siRNA-depleted for either Treacle (Treacle kd) or TOPBP1 (TOPBP1 kd) were treated with DMSO or 90 μM VP16 for 30 min. ChIP experiments were performed with antibodies against phospho-ATR (pATR; Ser1989), phospho-ATM (pATM; Ser1981) or γH2AX antibodies. ChIP was followed as described in (E). (G) HeLa cells were processed as described in (E). At least 2×106 sorted cells were used for ChIP with phospho-ATR (pATR; Ser1989), phospho-ATM (pATM; Ser1981) or γH2AX antibodies. ChIP was followed by qPCR using the d1 primers to the promoter of the rRNA gene (positioned as indicated on the scheme). Data are represented relative to the input. Values are means ±SD from at least three independent replicates.

Next, we investigated whether the interaction between Treacle and TOPBP1 in Treacle kd cells could be restored by the overexpression of different Treacle variants. The results of immunocytochemical analysis and chromatin immunoprecipitation with antibodies against TOPBP1 indicated that the expression of full-length Treacle in Treacle kd cells effectively restored the VP16-induced interaction between Treacle and TOPBP1, as well as the enrichment of TOPBP1 at the promoters of ribosomal genes. However, the expression of LLPS-defective forms of Treacle (Δ83-1121 or CS) did not lead to such restoration (Fig. 5D,E).

The response to DNA damage involves the activation of various signaling kinases and the recruitment of different repair factors to the DNA break site. Using chromatin immunoprecipitation followed by quantitative PCR (ChIP-qPCR) with antibodies against a panel of DNA repair proteins, it was confirmed that VP16 indeed induced DDR at the promoters of rDNA, including the activation of ATM/ATR kinases, phosphorylation of histone H2AX, and recruitment of repair factors such as 53BP1 and BRCA1 (Fig. 5F, S5C). As expected, nucleolar DDR was fully controlled by the interaction between Treacle and TOPBP1, as the knockdown of either of these proteins led to a dramatic reduction in VP16-induced repair signaling at rDNA (Fig. 5F, S5B,C). Finally, it was confirmed that nucleolar DDR could be efficiently restored in Treacle kd cells by overexpressing full-length Treacle, but not its LLPS-defect mutants (Δ83-1121 or CS) (Fig. 5G, S5D). Thus, we can conclude that the phase separation by Treacle plays a crucial role in facilitating its binding to TOPBP1 and activating DDR in conditions of genotoxic damage to rDNA.

Discussion

The nucleolus is a multicomponent phase condensate formed on the platform of rRNA, which is processed and incorporated into assembling ribosomes simultaneously with migration from FCs to DFCs and further to GC. Each of the nucleolar compartments provides a special environment formed by the demixing of a particular protein that constitutes a permanent component of a given compartment. Essential client proteins are subsequently recruited to this phase condensate. The constitutive components of DFC and GC are nucleolin and nucleophosmin (B23), respectively (Feric et al, 2016; Lafontaine et al, 2021). The debate continues regarding which protein plays a comparable role in FCs. The results of our study suggest that the phosphoprotein Treacle serves as the scaffold protein for nucleolar fibrillar centers in human cells. This aligns entirely with the findings of the E. Calo group, demonstrating that the evolutionary emergence of Treacle correlates with the appearance of the three-partite nucleolus in amniotes (Jaberi-Lashkari et al, 2023). They demonstrated that the introduction of human Treacle into zebrafish, which lack a competent ortholog for FC formation, induced the reorganization of the nucleolar structure from bipartite to tripartite, accompanied by the formation of FC-like structures.

Here, we demonstrated that Treacle organizes nucleolar FCs through its LLPS properties and also investigated the underlying mechanism leading to the phase separation of Treacle. The phase behavior of polymers is determined by interactions through associative motifs, referred to as stickers, separated by spacers, which are not the primary driving forces for phase separation (Choi et al, 2020). In Treacle, stickers consist of K/R and E/D residues in repeated disordered regions (RD) which generate alternating charged blocks. Meanwhile, a large number of glycine, proline, and glutamine residues, well-known as disorder-promoting residues (Romero et al, 2001; Theillet et al, 2013), can serve as effective spacers.

While the central domain of Treacle is responsible for its liquid behavior, this domain alone is unable to undergo autonomous phase separation within the cell. To undergo phase separation, the central domain requires a positively charged C-terminal domain, acting as a platform for nucleation. The mechanism by which the C-terminal domain facilitates nucleation is not entirely clear. We hypothesize that positively charged K residues in the C-terminal domain may interact with negatively charged E/D residues in RD, utilizing electrostatic interactions. Interestingly, the zebrafish ortholog of Treacle is homologous to human Treacle only in the central domain, lacking a homologous sequence for the C-terminal domain. This likely accounts for its inability to condense and form fully functional FC.

Endogenous Treacle may undergo post-translational modifications. Specifically, the central domain of Treacle is known to be heavily phosphorylated and ubiquitinated (Mooser et al, 2020; Werner et al, 2018; Werner et al, 2015). Currently, the role of these modifications is primarily attributed to facilitating interactions with partner proteins such as CUL3KBTBD8 or NOLC1 (Werner et al, 2018). However, it is known that phosphorylation can modulate phase separation, particularly by enhancing or reducing the polyampholytic properties of disordered regions (Yamazaki et al, 2022). As LLPS of Treacle relies on its polyampholytic nature, we do not exclude that post-translational ubiquitination and phosphorylation may alter charge ratios in its central domain, thereby modulating ampholytic properties and finely tuning phase separation.

The synthesis of 47S pre-rRNA occurs at the FC/DFC boundary, while primary processing initiates in the DFC. DFC formation is mediated by phase separation of fibrillarin gathered on pre-rRNA scaffold and is maintained as long as there is a directed radial flow of rRNA from FC to DFC (Yao et al, 2019). The role of Treacle in ensuring rRNA processing has been repeatedly described in the literature (Calo et al, 2018; Gonzales et al, 2005), but the mechanism of this phenomenon remained unclear. We demonstrated that the efficiency of rRNA processing is compromised not only with the complete depletion of endogenous Treacle but also with the expression of LLPS-defective mutants of Treacle. This led us to propose that Treacle facilitates rRNA processing not only through the direct recruitment of specific factors but also by organizing the cooperative structure of FC/DFC and separating FC components from DFC. Disruption and mixing of FC and DFC components due to Treacle’s impaired LLPS properties may interfere with the radial flow of rRNA, thus hindering its normal processing.

Additionally, we explored the role of Treacle’s phase separation in nucleolar DDR. Treacle’s well-known partner, TOPBP1, plays a crucial role in the activation of ATR during DNA replication stress (Day et al, 2021). The interaction between TOPBP1 and Treacle is essential for nucleoli-specific DDR and replication stress response (Mooser et al, 2020; Velichko et al, 2021; Velichko et al, 2019). This interaction occurs via phosphoserines 1227/1228 in a conserved acidic motif of C-terminal domain of Treacle (Mooser et al, 2020). However, we showed that Treacle with intact phosphoserines 1227/1228 but unable to undergo LLPS loses the ability to interact with TOPBP1 during genotoxic stress. This implies that for interaction with TOPBP1 and subsequent activation of nucleolar DDR, Treacle must also possess corresponding phase separation characteristics.

In conclusion, our study unveils Treacle’s role not merely as a structural scaffold for FC/DFC but as a nucleolar hub protein that integrates functions in ribosomal gene transcription, rRNA processing, and rDNA repair.

Materials and methods

Cell culture and drug treatment

Human HeLa (ATCC® CCL-2™) were cultured in DMEM (PanEco) supplemented with 10% fetal bovine serum (FBS; HyClone/GE Healthcare) and penicillin/streptomycin. The cells were cultured at 37°C in a conventional humidified CO2 incubator. DNA damage was induced by the treatment of cells with 90 µM etoposide (Sigma-Aldrich, #E1383) for 30 min or 1 h. For rRNA transcription inhibition, cells were treated with 0,05 µg/ml actinomycin D (Sigma-Aldrich, #A1410) or 20 µM CX-5461 for 3 h. To obtain 1,6-hexanediol (HD)-treated cells, HeLa cells were incubated with 5% 1,6-HD (Sigma-Aldrich, #240117) in serum-free medium at 37°C in a humidified atmosphere for 10 min. To obtain ammonium acetate-treated cells, HeLa cells were incubated with 200 mM ammonium acetate in a complete culture medium at room temperature for 5 min.

Plasmid constructs

For the FusionRED-Treacle-Cry2 (opto-Treacle) construct, the full-length Treacle was amplified by PCR from cDNA with primer set #1 (Supplementary Table 1) using KAPA High-Fidelity DNA Polymerase (KAPA Biosystems, KE2502). The forward and reverse primers contained XhoI sites. The amplified fragment was inserted into the FusionRed-C vector (Evrogen, FP411) linearized with XhoI. The Cry2 fragment was amplified by PCR from the plasmid pHR-mCh-Cry2WT (Addgene, #101221) with primer set #2 (Supplementary Table 1) digested with NheI and inserted into pFusionRed-Treacle linearized with NheI using NheI enzyme.

For the FusionRED-FUS-Cry2 construct, the FUS was amplified by PCR from the plasmid pHR-FUSN-mCh-Cry2WT (Addgene, #101223) with primer set #3 (Supplementary Table 1) using KAPA High-Fidelity DNA Polymerase (KAPA Biosystems, KE2502). The forward and reverse primers contained BspEI and KpnII sites respectively. The amplified fragment was inserted into the pFusionRed-C vector (Evrogen, FP411). The Cry2 fragment was amplified by PCR from the plasmid pHR-mCh-Cry2WT (Addgene, #101221) with primer set #2 (Supplementary Table 1) digested with NheI and inserted into pFusionRed-Treacle using NheI enzyme.

To generate Treacle-GFP or Treacle-Katushka2S constructs, the full-length Treacle was amplified by PCR from cDNA with primer set #4 (Supplementary Table 1) using KAPA High-Fidelity DNA Polymerase (KAPA Biosystems, KE2502). The forward and reverse primers contained BglII and BamHI sites respectively. The amplified fragment was inserted into the pTurboGFP-C (Evrogen, FP511) or pKatushka2S-C (Evrogen, FP761) vectors using BglII/BamHI restriction/ligation.

Treacle Δ1-83 deletion mutant was constructed based on the pTreacle-GFP full or pTreacle-2S full plasmid using iProof High-Fidelity DNA polymerase with primer set #5 (Supplementary Table 1). The resulting DNA template after PCR was reprecipitated and treated with the DpnI restriction enzyme. In the next step, the desired DNA template was purified on an agarose gel, phosphorylated with T4 Polynucleotide Kinase (T4 PNK), and ligated.

To obtain the charge-scrambled Treacle-GFP or Treacle-2S (Treacle-GFP CS or Treacle-2S CS), a fragment encoding 137-1130 aa of Treacle was removed from the pTreacle-GFP or pTreacle-2S plasmid using EcoR1 and HindIII restrictases. This fragment was then replaced with a synthetic sequence encoding the amino acid sequence with the necessary changes (see Supplementary Table 2).

Treacle Δ83-1121 deletion mutant was constructed based on the pTreacle-GFP full or pTreacle-2S full plasmid using iProof High-Fidelity DNA polymerase with primer set #6 (Supplementary Table 1). The resulting DNA template after PCR was reprecipitated and treated with the DpnI restriction enzyme. In the next step, the desired DNA template was purified on an agarose gel, phosphorylated with T4 Polynucleotide Kinase (T4 PNK), and ligated.

Treacle Δ1121-1488-NLS mutant was constructed based on the pTreacle-GFP full or pTreacle-2S full plasmid using iProof High-Fidelity DNA polymerase with primer set #7 (Supplementary Table 1). The resulting DNA template after PCR was reprecipitated and treated with the DpnI restriction enzyme. In the next step, the desired DNA template was purified on an agarose gel, phosphorylated with T4 Polynucleotide Kinase (T4 PNK), and ligated.

For the generation of Treacle ΔNOLC (Δ1350-1488) mutant, fragments were amplified by PCR from the pTreacle-GFP plasmid with primer set #8 (Supplementary Table 1). The forward and reverse primers contained BglII and BamHI sites respectively. The amplified fragment was inserted into either the pTurboGFP-C (Evrogen, FP511) or pKatushka2S-C (Evrogen, FP761) vectors using BglII/BamHI restriction/ligation.

Gene knockdown

RNA interference experiments were performed using Lipofectamine 3000 transfection reagent (Thermo Scientific) following the manufacturer’s instructions. The cells were transfected with 50 nM Treacle/TCOF1 siRNA (the sequences of the siRNA is provided in Supplementary Table S3) or 50 nM TOPBP1 (Santa Cruz Biotechnology, #sc-41068).

For CRISPR/Cas9-mediated knockout, two single guide RNAs (sgRNA) to first exon of the TCOF1 gene were designed using the guide RNA design tool (www.atum.bio/eCommerce/cas9/input). The sgRNA targeting sequences were separately cloned into the px330mCherry (Addgene #98750). A list of all oligonucleotides is provided in Supplementary Table S3. The plasmids were co-transfected into HeLa cells with LTX transfection reagent (Invitrogen). 5-7 days after transfection cells were fixed and immunostained with required antibodies.

Fluorescence microscopy

For immunostaining, cells were grown on microscope slides. All samples were fixed in 1% formaldehyde in PBS for 15 min at room temperature and treated with 1% Triton X-100 for permeabilization. Cells were washed in PBS and then incubated with antibodies in PBS supplemented with 1% BSA and 0.05% Tween 20 for 1 h at room temperature or overnight at 4°C. Then the cells were washed with PBS three times (5 min each time). The primary antibodies bound to antigens were visualized using Alexa Fluor 488-conjugated secondary antibodies. The DNA was counterstained with the fluorescent dye 4,6-diamino-2-phenylindole (DAPI) for 10 min at room temperature. The samples were mounted using Dako fluorescent mounting medium (Life Technologies). The immunostained samples were analyzed using a Zeiss AxioScope A.1 fluorescence microscope (objectives: Zeiss N-Achroplan 40 ×/0.65 and EC Plan-Neofluar 100 ×/1.3 oil; camera: Zeiss AxioCam MRm; acquisition software: Zeiss AxioVision Rel. 4.8.2; Jena, Germany) or STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2). The images were processed using ImageJ software (version 1.44) and analyzed using CellProfiler software (version 3.1.5). 3D reconstruction of xyz confocal datasets (z-stacks) was performed using Leica LAS-X software. Antibodies used in the study are listed in Supplementary Table 7.

Cell sorting

Cells were transiently transfected with required plasmids using LTX transfection reagent (Invitrogen) or immunostained with required antibodies. Cell sorting was performed using an SH800 Cell Sorter (Sony) with a laser tuned to 488 nm for green fluorescence and 561 nm for red fluorescence. Gates were set with reference to negative controls. A minimum of 2×106 events was collected for ChIP or RNA extraction.

RNA extraction and RT-qPCR

After sorting, living or immunostained cells were pelleted by centrifugation at 3000 g for 5′ at 4°C. The supernatant was discarded. Total RNA was isolated from the pellet as described (Hrvatin et al, 2014). All RNA samples were further treated with DNase I (Thermo Scientific) to remove the residual DNA. RNA (1 µg) was reverse transcribed in a total volume of 20 µl for 1 h at 42°C using 0.4 µg of random hexamer primers and 200 U of reverse transcriptase (Thermo Scientific) in the presence of 20 U of ribonuclease inhibitor (Thermo Scientific). The cDNA obtained was analyzed by quantitative polymerase chain reaction (qPCR) using the CFX96 real-time PCR detection system (Bio-Rad Laboratories). The PCRs were performed in 20 µl reaction volumes that included 50 mM Tris–HCl (pH 8.6), 50 mM KCl, 1.5 mM MgCl2, 0.1% Tween-20, 0.5 µM of each primer, 0.2 mM of each dNTP, 0.6 µM EvaGreen (Biotium), 0.2 U of Hot Start Taq Polymerase (Sibenzyme) and 50 ng of cDNA. Primers used in the study are listed in Supplementary Table 4.

Chromatin immunoprecipitation (ChIP) and ChIP-seq analysis

Living cells were fixed for 15 min with 1% formaldehyde at room temperature, and crosslinking was quenched by adding 125 mM glycine for 5 min. Cell sorting was performed if needed. Cells were harvested in PBS, and nuclei were prepared by incubation in FL buffer (5 mM PIPES, pH 8.0, 85 mM KCl, 0.5% NP40) supplemented with Protease Inhibitor Cocktail (Bimake) and Phosphatase Inhibitor Cocktail (Bimake) for 30 min on ice. Next, chromatin was sonicated in RIPA buffer (10 mM Tris–HCl, pH 8.0, 140 mM NaCl, 1% Triton X-100, 0.1% sodium deoxycholate, 0.1% SDS) with a VirSonic 100 to an average length of 200-500 bp. Per ChIP reaction, 10–20 µg chromatin was incubated with 2–4 µg antibodies overnight at 4°C. The next day, Protein A/G Magnetic Beads (Thermo Scientific) were added to each sample and incubated for 4 h at 4°C. Immobilized complexes were washed two times for 10 min at 4°C in low salt buffer (20 mM Tris-HCl, pH 8.0, 150 mM NaCl, 2 mM EDTA, 0.1% SDS, 1% Triton X-100) and high salt buffer (20 mM Tris–HCl, pH 8.0, 500 mM NaCl, 2 mM EDTA, 0.1% SDS, 1% Triton X-100). Samples were incubated with RNase A (Thermo Scientific) for 30 min at room temperature. The DNA was eluted from the beads and de-crosslinked by proteinase K digestion for 4 h at 55°C and subsequent incubation at 65°C for 12 h. Next, DNA was purified using phenol/chloroform extraction and analyzed by qPCR. The qPCR primers used for ChIP analysis are listed in Supplementary Table 5.

The sequencing libraries were then prepared with NEBNext Ultra II kit according manufacturer’s protocol. Final libraries were PCR amplificated and adapter dimers were cleaned with 1:1 MagPure magnetic beads (Magen Biotechnology). Resulted DNA was resuspended in 30 mkl 10 mM Tris-HCl buffer ph 8.0 and were sequenced on Illumina machine. Chip-seq reads were mapped to the reference human genome hg38 assembly using Bowtie v2.2.3 with the ‘–very-sensitive’ mode. Non-uniquely mapped reads, possible PCR and optical duplicates were filtered out using SAMtools v1.5. The bigWig files with the ratio of RPKM normalized ChIP-seq signal to the input were generated using deepTools v3.4.2 bamCompare function.

Cool files was generated, merged and normalized using cooler version 0.8.11 (https://github.com/open2c/cooler).

Electron microscopy

Twenty-four hours after transfection, cells were fixed with 2.5% neutralized glutaraldehyde in the requisite buffer for 2 h at room temperature, post-fixed with 1% aqueous OsO4, and embedded in Epon. Sections of 100 nm thickness were cut and counterstained with uranyl acetate and lead citrate. Sections were examined and photographed with a JEM 1400 transmission electron microscope (JEOL, Japan) equipped with a QUEMESA bottom-mounted CCD-camera (Olympus SIS, Japan) and operated at 100 kV.

In vitro Treacle condensation

Recombinant human Treacle protein (#YHG52701, Atagenix) was diluted with LLPS buffer (20 mM Tris-HCl, pH 7.4, 1 mM EDTA, 100 mM KCl) to a final concentration of 10 μM protein. Treacle LLPS was examined in the absence and presence of 1-10% dextran as a crowding agent. To test the effects of NaCL or 1,6-hexandiol on Treacle LLPS, NaCL or 1,6-hexandiol was added to the LLPS buffer with 10% PEG at 500 mM or 10% concentrations, respectively. The mixtures were analyzed on the Zeiss AxioScope A.1 fluorescence microscope equipped for brightfield imaging.

Live-cell imaging

Cells were seeded in 35-mm glass-bottom dishes twelve hours before transfection with the plasmid. Twenty-four hours after transfection, Hoechst 33342 (Cell Signaling Technology) was added to the medium at a final concentration of 1 μg/ml, and the cells were incubated for 20 min at 37°C. Hoechst-containing medium was replaced with a complete fresh medium before the cells underwent live-cell imaging using a STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2) equipped with an incubation chamber to provide a humidified atmosphere at 37°C with 5% CO2. For observation of liquid droplet behavior, z-stack time-lapses were taken. Maximum intensity projections are shown, but all findings were confirmed on a single plane.

Optodroplet assay

Cells were seeded in 35-mm glass-bottom dishes and were transfected with Treacle-FusionRed-Cry2 plasmid using LTX transfection reagent (Invitrogen) 24 h before imaging. Twenty-four hours after transfection, Hoechst 33342 was added to the medium at a final concentration of 1 μg/ml, and the cells were incubated for 10 min at 37°C. Hoechst-containing medium was replaced with a complete fresh medium before the cells underwent live-cell imaging using a STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2) equipped with an incubation chamber to provide a humidified atmosphere at 37°C with 5% CO2. Droplet formation was induced with light pulses at 488 nm (blue light, 1% laser power) for 10 sec, and z-stack images were captured every 60 sec in the absence of blue light.

FRAP analysis

Samples for FRAP were prepared as described above for live-cell imaging experiments. FRAP experiments were performed on STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2) equipped with an incubation chamber to provide a humidified atmosphere at 37°C with 5% CO2. Regions of interest were bleached using 405-nm laser (1% laser power).

The measurements involved 3 prebleaching frames, one flashes of bleaching (1% of laser power), and 28 postbleaching frames (5 s/frame). Individual FRAP traces were normalized to maximal prebleach and minimal postbleach intensities.

Proximity ligation assay

After drug treatments or at 24 h post-transfection, cells were fixed for 15 min with 1% formaldehyde at room temperature and permeabilized with 1% Triton X-100. Next, cells were subjected to the proximity ligation assay (PLA) using the Duolink Green (Sigma-Aldrich) according to the manufacturer’s instructions. Briefly, cells were blocked, incubated with appropriate primary antibodies overnight at 4°C, and then incubated with PLA probes, which are secondary antibodies (anti-mouse-IgG and anti-rabbit-IgG) conjugated to unique oligonucleotides. Samples were then treated with a ligation solution followed by an amplification solution containing polymerase and fluorescently labeled oligonucleotides, allowing rolling-circle amplification and detection of discrete fluorescent dots. Some samples after the PLA protocol were additionally counterstained with Alexa-Fluor-488- or Alexa-Fluor-594-conjugated (Molecular Probes) secondary antibody. The DNA was counterstained with DAPI for 10 min at room temperature. The samples were mounted using Dako fluorescent mounting medium (Life Technologies) and analyzed using STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2). The images were analyzed using ImageJ software (version 1.44).

Transcription labeling

For EU incorporation, the cells were incubated with 100 μM EU (Sigma-Aldrich) for 2 hr at 37°C. Then, the cells were washed three times with PBS and fixed in ice-cold methanol for 10 min. The samples were then processed using a Click-iT EU Imaging Kit (Life Technologies) according to the manufacturer’s recommendations.

Single molecule RNA Fluorescentin in situ Hybridization (smFISH)

All single-molecule RNA FISH probes were designed as described (Yao et al, 2019) and labeled with FITS on the 3’ends (Supplementary Table 6). Cells were fixed with 4% PFA for 15 min, followed by permeabilization with 1% Triton X-100 for 10 min. Cells were incubated in 10% formamide/2x SSC for 10 min at room temperature and then were hybridizated with 5 nM each of RNA probes in 50% formamide/2x SSC at 37◦C for 16 hours. After hybridization, the cells were washed 10% formamide/2x SSC for 30 min at 37◦C.

Supporting information

eLife Assessment

In this important study, the authors provide evidence that Treacle, a disease-relevant intrinsically disordered protein, undergoes biomolecular condensation to support the structure and function of the fibrillar center of the nucleolus. The findings, arising from complementary approaches, provide solid evidence for the role of Treacle condensation in supporting rDNA transcription, rRNA processing, and genome integrity. These findings may be of interest to the communities studying biomolecular condensates, nucleolar organization, and ribosome biogenesis.

Reviewer #1 (Public Review):

Summary:

The manuscript by Velichko et al. argues that the ability of nucleolar protein Treacles to form phase-separated condensates is necessary for its function in nucleolar organization, rRNA transcription, and rDNA repair. These findings may be of interest to the communities studying biomolecular condensates, nucleolar organization, and ribosome biogenesis. The authors propose that Treacle's ability to undergo liquid-liquid phase separation is the key to its role as a scaffold for the FC of the nucleolus. The experiments in this study were designed and performed well, particularly the overexpression studies, done in the absence of endogenous protein and accounted for the protein expression levels. However, in my view, the interpretation of these data should consider the possibility that specific protein-protein interactions of Treacle may also play a role in the organization of the FC compartment in vivo. The in vivo results do not exclude, and sometimes imply the presence of specific protein-protein interactions that may drive the organization of FC instead of, or in addition to LLPS.

Main points:

In the first part of the manuscript, the depletion of Treacle disrupted the FC and its (somewhat arbitrary) boundary with the dense fibrillar component, as well as rRNA biogenesis. The phenotypic effects of Treacle depletion by gene knockout or siRNA knockdown were evaluated thoroughly, and I see no issues here except that all experiments were conducted in HeLa cells, and it may not hurt to validate some key findings in a more normal cell line.

Next, the authors tested the hypothesis that the function of Treacle is due to its ability to form biomolecular condensates. In vitro, recombinant Treacle displayed classical phase separation behavior, forming liquid droplets at low salt concentrations and in the presence of dextran. Similarly, overexpression of fluorescently tagged Treacle at high concentrations showed classical liquid droplet behavior, characterized by round shapes and rapid fusion, which is illustrated by beautiful live cell video microscopy. The issue I see here is with the interpretation: the formation of classical phase-separated droplets at high concentrations suggests that Treacle may require reaching a certain saturating concentration to undergo phase separation. In other words, high levels of overexpressed protein might lead to abnormal phase separation that may not happen under normal expression levels. Based on these results, it is not necessarily correct to assume that its normal conformation is solely due to phase separation, as the formation of condensates at saturating concentrations does not automatically imply that the same components undergo phase separation under physiological conditions.

Treacle had been previously reported to interact with other proteins, specifically RPA194 and UBF, and these interactions were mapped to specific domains: the central repeated domain reportedly binds to RNA Pol I, while the C-terminus is involved in rDNA promoter recognition and UBF recruitment. Both of these proteins are necessary for rRNA transcription and nucleolar formation. Authors showed that overexpressing mutants impaired in phase separation resulted in defects in ribosomal RNA transcription and processing, as well as reduced DNA damage response efficiency. Specific protein-protein interactions as potential drivers of compartmentalization should be factored into the interpretation of these results. For instance, the deletion of the C-terminal (Δ1121-1488) results may indicate that the interaction with UBF is important. A charge-scrambled central domain mutant may have lost its interaction with Pol I. These specific interactions may establish the architecture of the compartment and increase the local concentration of Treacle, which in turn could facilitate phase separation locally. LLPS and specific protein-protein interactions are not mutually exclusive.

Overall, the data supports the idea that the overexpressed Treacle behaves like a classic phase-separated protein, but it is still possible that at physiological levels its specific interactions with other proteins are also important for the organization of FC. I am not suggesting that authors performed a conceptually different work, but this aspect should be discussed in the manuscript.

Other points:

FACS - sorting used throughout the study to separate treatment from the control essentially distinguishes transfected vs untransfected cells. Since the transfection itself can have odd effects, it might be beneficial to include an additional control involving Cas9 transfection with a non-targeting guide RNA.

The authors convincingly demonstrated in Figure 1 that the depletion of Treacle reduces RPA194 occupancy on the rDNA. This raises a question: which Treacle mutants can restore RPA194 occupancy, and which cannot?

Figure 2 - measuring FRAP recovery rates as indicative of LLPS, at least for the full-length Treacle, would be more informative if authors assessed the protein turnover within the compartment (half or partial FRAP) versus exchange in and out of the compartment (full compartment FRAP).

Statement related to Figure 2: "Fluorescence recovery in FCs, nucleolar caps, and extranucleolar condensates never reached the initial values over the analyzed time periods. This suggests that the high molecular exchange rate occurs through the mixing of Treacle molecules within the condensate boundaries and does not involve external diffusion". Assuming the post-bleach data were normalized to the cell's total fluorescent intensity, the presence of a substantial immobile fraction could also suggest high-affinity binding of that fraction to something within the compartment.

Data related to DDR activation in ribosomal genes under genotoxic stress (Figure 5) is convincing, but it would not hurt to confirm the key findings in a more normal cell line, since HeLa cells may not accurately represent all aspects of healthy DDR.

Reviewer #2 (Public Review):

Summary:

Velichko, Artem, et al. investigate the role played by the long intrinsically disordered protein Trecle in nucleolar morphology and function, with an interest in its potential ability to undergo liquid-liquid phase separation. The authors explore Treacle's role in core functions of the nucleolus (rRNA biogenesis and DNA repair), which has been a subject of continual investigation since it was identified that truncation of Treacle is the major genetic cause of Treacher-Collins syndrome. They show that knock out of Treacle leads to de-mixing of canonical markers of the FC (UBF, RPA194) and DFC (FBL) phases of the nucleolus. They also show that replacing Treacle with mutants that disrupt its bulk dynamics leads to the de-mixing of FBL. These mutants either remove the central region of Treacle (∆83-1121) or, more subtly, reduce the segregation of charged residues by scrambling them (CS- Charge Scrambled). The observed morphological disruptions mirrored disruptions to the production of rRNA and the ability to recruit the DNA-damage response factor TOPBP1. These data give new insight into the role played by the central region of Treacle in affecting its bulk dynamics and the potential effects of disruptions therein to nucleolar morphology and function.

Strengths:

The characterizations of changes to nuclear morphology upon Treacle knockout is the major strength of this study (Figure 1). Methodologically the CRISPR knockout appears sound. The characterized effects on the canonical markers of the FC and DFC phases support the idea that Treacle has a scaffolding function. While the effect of Treacle perturbations has been studied before, this has often been phenotyped in the context of development or rRNA biogenesis, and less often on the sub-cellular level.

The other major strength of this study is its characterization of the effects of the charge scramble mutant. The authors find that replacing endogenous Treacle with this mutant reduces the bulk dynamics of Treacle (Figure 3K-M), de-mixes FBL from the DFC (Figure 4C-D), lowers pre-rRNA synthesis (Figure 4E-G), and abolishes the recruitment of the DNA-damage response factor TOPBP1 (Figure 5).

Weaknesses:

Clarity around the reagents used and deeper analyses would bolster the author's claims about the condensation behavior of Treacle.

Limited characterization and sparse methodological details regarding recombinant Treacle lead to a concern about the observation that Treacle condenses in vitro. The concerns are offset by the fact that most of the paper uses cellular data to draw conclusions.

The authors ascribe liquid-like behavior to Treacle based on spherical morphology and fusion events of Treacle-Katushka2S condensates as well as fluorescence recovery after a photobleaching (FRAP); these are accepted characterizations in the biomedical field. Nonetheless, the authors only use FRAP to characterize mutants, which limits conclusions about their apparent material state. Overall, FRAP data are better interpreted as a readout of bulk dynamics. For example, the FRAP traces of Treacle plateau at a recovery percentage between 40 and 60%, indicating complex bulk dynamics and the possibility of an immobile pool that is not liquid-like.

Lastly, the Treacle-Katushka2S construct is the predominant construct used throughout the paper. The known tetrameric nature of Katushka2S contrasts with the presumptively monomeric Treacle-FusionRed-Cry2 construct. This is relevant because multi-valance is known to increase the driving forces for condensation and affect condensate material properties. The authors report that the Treacle-FusionRed-Cry2 construct (monomeric) exhibits less condensation than the Treacle-Katushka2S construct (tetrameric). Thus, one is left concerned that the latter construct is not wholly representative of intrinsic Treacle condensation behavior.

Reviewer #3 (Public Review):

Summary:

This study provides evidence that the protein Treacle plays an essential role in the structure and function of the fibrillar center (FC) of the nucleolus, which is surrounded by the dense fibrillar component (DFC) and the granular component (GC). The authors provide new evidence that, like the DFC and GC, the functional FC compartment involves a biomolecular condensate that contains Treacle as a key component. Treacle is essential to the transcription of the rDNA as well as proper rRNA processing that the authors tie to a role in maintaining the separation of FC components from the DFC. In vitro and in vivo experiments highlight that Treacle is itself capable of undergoing condensation in a manner that depends on concentration and charge-charge interactions but is not affected by 1,6 hexanediol, which disrupts weak hydrophobic interactions. Attempting to generate separation-of-function mutants, the authors provide further evidence of complex interactions that drive proper condensation in the FC mediated by both the central repeat (low-complexity, likely driving the condensation) and C-terminal domain (which appears to target the specificity of the condensation to the proper location). Using mutant forms of Treacle defective in condensation, the authors provide evidence that these same protein forms are also disrupted in supporting Treacle's functions in rDNA transcription and rRNA processing. Last, the authors suggest that cells lacking Treacle are defective in the DNA damage response at the rDNA in response to VP16.

Strengths:

In general, the data are of high quality, the experiments are well-designed and the findings are mostly carefully interpreted. The findings of the work complement prior high-impact studies of the DFC and GC that have identified constituent proteins as the lynchpins of the biomolecular condensates that organize the nucleolus into its canonical three concentric compartment structure and are therefore likely to be of broad interest. The attempts to generate separation-of-function mutants to dissect the contribution of condensation to Treacle function are ambitious and critical to demonstrating the relevance of this property to the biology of the FC. The complementarity of the methods applied to investigate the Treacle function is appropriate and the findings integrate well towards a compelling narrative.

Weaknesses:

Although the attempt to generate separation of function mutants of Treacle is laudable (and essential), there still remain possible alternative explanations for the observed defects in such mutants as most of the experimental approaches give rise to negative results. The DDR angle of the manuscript seems somewhat more preliminary as it is largely restricted to looking at the recruitment of DDR factors to the rDNA in response to a specific insult (VP16). It would be more compelling if the authors could investigate a more biologically relevant outcome (e.g. rDNA repeat number stability).

Author response:

The following is the authors’ response to the original reviews

Recommendations for the authors:

Reviewer #1 (Recommendations for the Authors):

The interpretation of results obtained with opto-Treacle (related to Figure 2C) may be expanded.

We thank the reviewer for their insightful comment regarding the interpretation of the results obtained with opto-Treacle. We understand the concern that the difference in the size of the condensates formed by opto-Treacle (Figure 2C) compared to Treacle-2S or other constructs may raise questions about the role of tetramerization in driving condensate formation, as 2S is known to tetramerize while FusionRed is not susceptible to multimerization.

To address this concern, we emphasize that we have demonstrated that overexpressed Treacle forms large condensates even in the absence of any fluorescent protein, as included in the revised manuscript. This observation supports the conclusion that Treacle's ability to form condensates is intrinsic and does not depend on the multimerization capacity of the fluorescent tag.

We believe that the observed difference in condensate size between opto-Treacle and Treacle-2S, Treacle-GFP, or untagged Treacle arises primarily from the time available for condensate assembly. Opto-Treacle condensation occurs rapidly, within approximately 10 seconds of blue light illumination, whereas Treacle-2S, Treacle-GFP, or untagged Treacle undergo condensation over the extended period of 24–48 hours of protein overexpression. This temporal difference likely accounts for the disparity in condensate size, as longer assembly times allow for larger and more mature condensates to form.

Given this reasoning, we consider it unnecessary to further emphasize the size differences in the main text of the article, as we believe the underlying explanation is clear and supported by the data. Nonetheless, we are open to incorporating additional clarifications if the reviewer deems it necessary.

The authors might reconsider referring to Treacle as a scaffold. Ultimately, the scaffold for the nucleolus is the rDNA with its bound proteins. Scaffold proteins, by definition, bind multiple protein partners and facilitate the formation of multiprotein complexes, a role not really attributed to homotypic LLPS.

We thank the reviewer for raising this important point regarding the use of the term "scaffold" in relation to Treacle. We fully acknowledge that rDNA, along with its associated protein complexes, serves as the primary structural scaffold for the nucleolus. However, we believe that referring to Treacle as a scaffold is appropriate and justified within the specific context of our study.

First, we emphasize that we describe Treacle as a scaffold specifically for nucleolar fibrillar centers (FCs), rather than for the nucleolus as a whole. This distinction is important, as our work focuses on the role of Treacle in organizing FC components, rather than the broader structural organization of the nucleolus.

Second, as the reviewer notes, scaffold proteins are defined by their ability to bind multiple protein partners and facilitate the formation of multiprotein complexes. Our findings demonstrate that Treacle's condensation properties promote the binding and retention of key rDNA-associated protein partners, including RPA194, UBF, and Fibrillarin, within the FCs. This activity aligns with the functional definition of a scaffold protein, as Treacle supports the spatial organization and cooperative interactions of FC components essential for rRNA transcription and processing. Therefore, while we appreciate the reviewer's observation regarding the central role of rDNA as a nucleolar scaffold, we maintain that the use of the term "scaffold" to describe Treacle's role in organizing FCs is consistent with its demonstrated functional properties.

If authors decide to add the "Ideas and Speculation" subsection to their Discussion, it may be interesting to discuss the following outstanding questions: does Treacle undergo homotypic or heterotypic LLPS? Does its overexpression favor homotypic interactions? How does it segregate FC and DFC compartments -by exclusion? How does phase-separated Treacle interact with other proteins?

We thank the reviewer for these insightful questions. While we believe that adding a dedicated "Ideas and Speculation" subsection would be redundant, we have already addressed the questions regarding Treacle’s homotypic or heterotypic LLPS and its interactions with other proteins in the revised "Discussion" section. Additionally, we have included a new section in the manuscript specifically focused on investigating the role of Treacle condensation in its interactions with protein partners, further expanding on these points.

In Materials and Methods, smFISH section -"probes were designed as described (Yao et al, 2019) and labeled with FITS on the 3'ends" - was it meant to say FITC (i.e. Fluorescein)?

We thank the reviewer for catching this error. This was indeed a typo, and we have corrected it to "FITC (i.e., Fluorescein)" in the revised text.

Reviewer #2 (Recommendations for the Authors):

Regarding recombinant Treacle, the main concern is that the authors may not be observing the condensation of Treacle itself. The quality of the purchased recombinant Treacle is unclear (this reviewer could not find Treacle listed on the vendor website despite using the supplied catalog number or vapors search terms). Furthermore, it is not clear if the condensates observed are Treacle or potentially the Dextran crowder. Only small percentages (>1%-5%) of either Dextran or PEG are needed to induce phase separation in two-component mixtures of these polymers. PEG may be in the Treacle storage butter. In addition to clarifying the State of recombinant Treacle, these concerns could be further assuaged by direct visualizing of Treacle forming condensates (via fluorescent n-terminal tagging) and filling in more of the phase space to observe the loss of condensates at a threshold concentration of Treacle. In general, the gold standard for establishing condensation of a given protein is mapping the full binodal phase diagram diagram of the protein. Understanding that protein is a limited resource, most groups simply map the lower concentration arm of the binodal, and this is sufficient to characterize a protein as having intrinsic condensation behavior. A similar mapping effort of Treacle would be welcomed.

We thank the reviewer for their thoughtful comments and for highlighting concerns regarding the interpretation of our experiments with commercial recombinant Treacle. We recognize the importance of ensuring that the observed condensation properties are intrinsic to Treacle and not influenced by potential contaminants, storage buffer components, or tags on the protein.

To address these concerns, we have re-evaluated the condensation properties of Treacle using a recombinant fragment independently purified in our laboratory. Specifically, we expressed and purified a Treacle fragment (amino acids 291–426), which includes two S/E-rich low-complexity regions (LCRs) and two linker regions, in E. coli. The protein was expressed as a TEV-cleavable maltose-binding protein (MBP) fusion, purified under native conditions via amylose resin, and subjected to TEV cleavage. This was followed by ion-exchange chromatography and extensive dialysis to remove any remaining impurities. These additional steps ensured that the purified Treacle fragment was of high purity and free from confounding components, such as polyethylene glycol (PEG). We have included detailed descriptions of this protocol in the revised manuscript.

Using this purified Treacle fragment, we confirmed its intrinsic condensation behavior in vitro. In the presence of 5% PEG8000 as a crowding agent, the fragment formed liquid-like condensates that exhibited spherical morphology and dynamic fusion events, key hallmarks of liquid-liquid phase separation (LLPS). Additionally, we demonstrated that the condensation of this Treacle fragment was sensitive to changes in pH and salt concentration but unaffected by 1,6-hexanediol treatment, suggesting that the condensates are stabilized predominantly by electrostatic interactions (Fig. 4B of the revised manuscript). Importantly, these findings provide robust evidence that Treacle possesses intrinsic phase-separation properties. All results from the commercial Treacle protein used in the initial version of the manuscript have been replaced with data obtained using this independently purified recombinant fragment.

We undestand that the condensation behavior of the fragment may not fully capture the behavior of full-length Treacle. Nevertheless, the in vitro experiments provide valuable mechanistic insights into the biophysical properties of Treacle. Furthermore, as emphasized in the revised manuscript, our study primarily focuses on understanding the condensation and functional role of Treacle in a cellular context, where we observe its critical involvement in organizing nucleolar structure and regulating rRNA transcription. These cellular experiments highlight the biological relevance of Treacle’s condensation behavior.

With regard to mapping the binodal phase diagram of Treacle, we concur with the reviewer that such an effort would be ideal for a more comprehensive characterization of Treacle’s condensation properties. However, the limited availability of purified protein currently precludes a detailed mapping effort. Despite this limitation, we believe the qualitative assessments of Treacle’s condensation under varying conditions, now included in the revised manuscript, sufficiently demonstrate its intrinsic ability to phase-separate.

In conclusion, we are grateful for the reviewer’s feedback, which has allowed us to refine our methodology and strengthen the evidence supporting the intrinsic condensation properties of Treacle. We are confident that the revised manuscript provides a robust and thorough characterization of Treacle’s phase-separation behavior and its functional role in the cell, addressing the reviewer’s concerns. Thank you for your constructive recommendations, which have significantly improved the quality of our work.

Replacing 'liquid-phase' and 'liquid' with 'liquid-like' would make the language consistent with other papers in the field and more accurately reflect the degree of material state analysis carried out in the study.

We thank the reviewer for this insightful recommendation. In response to the suggestion, we have revised the manuscript to replace the terms "liquid-phase" and "liquid" with "liquid-like" throughout the text. This change ensures consistency with terminology commonly used in the field and more accurately reflects the degree of material state analysis performed in our study. We believe this adjustment improves the clarity and precision of our findings, aligning the manuscript with standard practices in the field. Thank you for helping us enhance the quality of the presentation.

The 'unclear' nature of the condensation behavior of the FC phase of the nucleolus is listed as a motivation for carrying out the study in the introduction; the authors could note here two recent papers that have investigated the nature of FC condensation: Jaberi-Lashkari et al. 2023 and King et al. 2024. The reviewer notes that while these were both pre-printed in late 2022, they were only recently published.

We thank the reviewer for bringing these recent studies to our attention. In response to the suggestion, we have cited the papers by Jaberi-Lashkari et al. (2023) and King et al. (2024) in both the introduction and discussion sections of the revised manuscript. These references are highly relevant to the context of our study and provide valuable insights into the condensation behavior of the FC phase of the nucleolus. We agree that incorporating these works strengthens the framing of our study and situates it more effectively within the broader field. Thank you for this constructive recommendation.

The statement that Treacle is "the main molecule present in the FC" is a substantial claim that does not need to be made to promote the author's case, nor is it well supported by the provided reference (Gal et al., 2022).

We thank the reviewer for pointing out this overstatement in our original manuscript. In response, we have revised the text to provide a more accurate and well-supported description. Specifically, we have replaced the claim that Treacle is "the main molecule present in the FC" with a statement highlighting its direct interactions with UBF and RNA Pol I, as well as its colocalization with these proteins within the FC. This revision ensures alignment with the provided references and more accurately reflects the current understanding of Treacle's role in the FC. We appreciate the reviewer's attention to this detail, which has helped us improve the clarity and accuracy of our manuscript.

The statement that "Treacle is one of the most intrinsically disordered proteins" is vague and unnecessarily grand. Treacle is a fully intrinsically disordered protein; these comprise 5% of the human proteome (Tsang et al. 2020), so Treacle is, indeed, unusual in that regard.

We thank the reviewer for highlighting the vague and unnecessarily broad nature of the original statement. In response, we have revised the text to provide a more precise and accurate description of Treacle's structural properties. Specifically, we replaced the claim that "Treacle is one of the most intrinsically disordered proteins" with the statement that "According to protein structure predictors (e.g., AlphaFold, IUPred2, PONDR, and FuzDrop), Treacle is a fully intrinsically disordered protein." This wording reflects the unique nature of Treacle while remaining scientifically accurate and supported by reliable computational predictions. We appreciate the reviewer's feedback, which has allowed us to improve the rigor and clarity of our manuscript.

A comment on the implications of the immobile pool of Treacle (which appears to be ~50% in WT and across a range of mutants) would be welcome. Additionally, the limitations of FRAP for interrogating material properties of condensed material in living systems are provided in Goetz and Mahamid, 2020. In this paper, the authors review instances where the ultrastructure of condensate is known and where FRAP data is available. They show that crystalline assemblies can recover faster than apparently liquid, spherical assemblies. A comment in the text about how these limitations apply to this study would be welcome.

We appreciate the reviewer’s insightful comments regarding the interpretation of the immobile pool of Treacle and the limitations of FRAP for characterizing material properties in living systems. As noted in our response to the public review, we believe the ~50% recovery rate after photobleaching observed in our experiments is best explained by the redistribution of Treacle molecules within the condensate, rather than significant exchange with the surrounding phase. This interpretation is strongly supported by the full- and half-FRAP analyses included in the revised manuscript, which demonstrated internal mixing dynamics within the condensates.

There appears to be a typo in the following sentence: "The highly positively charged CD serves as the nucleation center for RD but exhibits ambivalent phase properties, transitioning from LLPS to LSPS in the absence of rRNA." The LLPS to LSPS behavior was observed for mutants to the central domain (RD), not the c-terminal domain (CD).

Throughout the authors report single snapshots of representative cells and single line traces. Analysis of the key morphological feature across the population of cells would help the reader understand how widespread the observed phenotype is.

We thank the reviewer for raising this important point regarding the representation of morphological features across the cell population. To address this concern, we have included widefield micrographs of cell fields in the revised figures to provide a more comprehensive view of the phenotypes observed.

The statement that "The phase behavior of polymers is determined by interactions through associative motifs, referred to as stickers, separated by spacers, which are not the primary driving forces for phase separation" could be improved by pointing out that this is potentially incomplete for describing the kind of condensation that highly charged polymers undergo. The high charge and charge segregation of Treacle suggest that it is a blocky polyampholyte and that it condenses by coacervation. Models of associative polymers can be useful for describing coacervation, however, the driving forces for coacervation are less understood and have been proposed to include an entropic component (see Sathyavageeswaran et al. 2024, Sing and Perry 2020 and work from their groups as well as the Obermayer (Columbia) and Terrell (U. Chicago) Groups).

We thank the reviewer for highlighting this important aspect of the phase behavior of charged polymers and for suggesting relevant references. In response, we have revised the discussion section of the manuscript to include a more nuanced explanation of the condensation mechanisms for highly charged polymers such as Treacle. Specifically, we now describe Treacle as a blocky polyampholyte, suggesting that its condensation behavior may be driven by coacervation mechanisms.The relevant references have been added to the discussion section of the revised manuscript.

In addition to the above, the authors may consider citing two recent publications from the Pappu group (King et al. Cell 2024 and King et al. Nucleus 2024) that directly investigate the condensation potential of K-rich and E/D-rich' grammars' on nucleolar proteins and show that, like the authors, the K-rich region is essential for localization and is conserved across nucleolar proteins.

We thank the reviewer for bringing these relevant publications to our attention. The suggested references from the Pappu group (King et al., Cell 2024, and King et al., Nucleus 2024) have been added to the introduction and discussion sections of the revised manuscript, and their findings have been appropriately integrated into our analysis.

The authors could consider replacing the use of LLPS with a more generic term such as "condensation" or "biomolecular condensation." LLPS of polymers is a segregative transition driven by its incompatibility with the surrounding solvent. As indicated, Treacle is likely to be undergoing some form of coacervation (which is predominantly an associative tradition), which can be genetically described as condensation. See Pappu et al. 2023 for more details.

We thank the reviewer for their insightful suggestion. Following the reviewer's recommendation, we have replaced the term "LLPS" with "condensation" or "coacervation" throughout the manuscript, where appropriate. Additionally, we have referenced Pappu et al. (2023) and other to provide further context and clarity regarding the distinctions between these terms.

The authors cite Yao et al. 2019, but do not cite the follow-up study (Wu et al. 2021) or provide a statement on how the Chan group finds a role for the RGG domain of FBL in keeping the certain canonical markers of the FC and DFC de-mixed.

We thank the reviewer for pointing out these important references. The relevant citations, including Wu et al. (2021), have been added to the manuscript.

Reviewer #3 (Recommendations for the Authors):

The following comment is true but could be broadened to include examples of structured regions promoting biomolecular condensation. "In biological systems, phase separation is mainly a characteristic of multivalent or intrinsically disordered proteins (Banani et al, 2017; Shin & Brangwynne,2017; Uversky, 2019)."

We have expanded the statement as recommended by the reviewer: "In biological systems, phase separation is facilitated by a combination of multivalent interactions mediated by intrinsically disordered proteins and site-specific interactions that drive percolation."

Related to Figure 1.

The authors report Treacle-dependent EU incorporation (Figure 1D), but are there any changes more broadly to nucleolar number or size as a consequence? How do the authors interpret that the quantitative effect of AMD treatment is more extreme than Treacle depletion (Figure 1E).

We thank the reviewer for raising these important points. Regarding nucleolar number and morphology, we did not observe a change in the number of nucleoli upon Treacle depletion. However, nucleoli appeared more regularly rounded under these conditions, which we interpret as a consequence of the decreased rDNA transcription activity caused by Treacle depletion. A similar rounding of nucleoli is also observed upon actinomycin D (AMD) treatment, which is consistent with reduced transcriptional activity.

As for the more pronounced effect of AMD compared to Treacle depletion on EU incorporation, this can be explained by the fundamentally different mechanisms through which these conditions affect transcription. Treacle depletion reduces the local concentration of transcription factors at rDNA sites, thereby impairing transcription initiation and elongation to a certain extent. However, under Treacle depletion, RNA polymerase I still retains the ability to bind to the promoter and support a residual level of transcription. In contrast, AMD acts as a potent intercalator in GC-rich regions of rDNA, physically blocking the ability of RNA polymerase I to move along rDNA, resulting in near-complete cessation of rRNA synthesis.

Related to Figure 2.

The authors observe that AMD leads to coalescence of individual Treacle-2S+ bodies (e.g. Figure 2E) - does this suggest that ongoing rRNA transcription is required to prevent such events?

Thank you for your thoughtful question. Indeed, our observations strongly suggest that ongoing rRNA transcription is required to prevent the coalescence of Treacle-2S+ bodies, as observed upon AMD treatment. This interpretation aligns with the findings of Tetsuya Yamamoto et al., who demonstrated that nascent ribosomal RNA (pre-rRNA) acts as a surfactant to suppress the growth and fusion of fibrillar centers (FCs) in the nucleolus. Their work highlighted that nucleolar condensates formed via liquid-liquid phase separation (LLPS) tend to grow to minimize surface energy, provided sufficient components are available. However, the transcription of prerRNA stabilizes FCs by maintaining multiple microphases, preventing coalescence unless transcription is inhibited.

According to Yamamoto et al., nascent pre-rRNAs tethered to FC surfaces by RNA Polymerase I generate lateral pressure that counteracts interfacial tensions, effectively suppressing FC fusion. This activity is analogous to the surfactant properties of molecules in physical systems. When transcription is inhibited (e.g., by AMD), the loss of nascent rRNA allows condensates to coalesce, consistent with the behavior we observe.

We further propose that the AMD-induced coalescence of Treacle-2S+ bodies reflects the loss of this surfactant-like effect, as transcriptional activity ceases. This theory is also supported by the observation that Treacle condensates in the nucleoplasm, where rRNA transcription is absent, form larger structures. Collectively, these insights highlight the critical role of ongoing rRNA transcription in maintaining the structural integrity and dynamic organization of nucleolar substructures.

Related to Figure 3.

In the figure panels B-H the DAPI signal in gray obscures the Treacle localization, especially in Figure 3H. A non-merged image for each of these examples for the Treacle localization would be very helpful.

We thank the reviewer for this observation. To address this, we have included wide-field images without the DAPI overlay for the deletion mutant lacking the 1121-1488 region. These are now presented in Supplementary Figure S5G of the revised manuscript.

Related to Figure 5.

Only a single representative nucleus is shown in the PLA analysis presented in Figure 5B.

Quantification to assess the robustness of this response with the addition of VP16 is needed. The authors use ChIP and immunocytochemistry as orthogonal methods but it would be best to therefore show both for each manipulation that is performed - the immunostaining of TOPBP1 in the Treacle KD cells in S5A should be in the main Figure 5 to complement transformation of constructs as in Figure 5D.

We appreciate the reviewer’s comment. To address this, we performed a quantitative analysis of PLA fluorescence signals in control and etoposide-treated cells, and the results are now presented in Supplementary Figure S8C. Additionally, as recommended, we have transferred the results of the immunocytochemistry of TOPBP1 in Treacle KD and Treacle KN cells to the main figure, now included as Figures 7D-E in the revised manuscript.

Version 2

Abstract

We investigated the role of the nucleolar protein Treacle in organizing and regulating the nucleolus in human cells. Our results support Treacle’s ability to form liquid-like phase condensates through electrostatic interactions among molecules. The formation of these biomolecular condensates is crucial for segregating nucleolar fibrillar centers from the dense fibrillar component and ensuring high levels of rRNA gene transcription and accurate rRNA processing. Both the central and C-terminal domains of Treacle are required to form liquid-like condensates. The initiation of phase separation is attributed to the C-terminal domain. The central domain is characterized by repeated stretches of alternatively charged amino-acid residues and is vital for condensate stability. Overexpression of mutant forms of Treacle that cannot form liquid-like phase condensates compromises the assembly of fibrillar centers, suppressing rRNA gene transcription and disrupting rRNA processing. These mutant forms also fail to recruit DNA topoisomerase II binding protein 1 (TOPBP1), suppressing the DNA damage response in the nucleolus.

Introduction

Applying polymer chemistry principles to biological molecules has significantly broadened our understanding of the functions of membraneless compartments (Mehta & Zhang, 2022; Mittag & Pappu, 2022). Multiple weak, cooperative, and dynamic interactions underlie the assembly of these self-organized structures, termed biomolecular condensates. Mechanistically, liquid–liquid phase separation (LLPS)—a demixing process that yields a condensed protein-enriched phase and a dilute phase—can mediate the self-assembly of biomolecular condensates. In biological systems, phase separation is facilitated by a combination of multivalent interactions mediated by intrinsically disordered regions and site-specific interactions that drive percolation (Banani et al., 2017; Shin & Brangwynne, 2017; Uversky, 2019). Biomolecular condensates serve as functional hubs in diverse cellular processes such as transcription (Sabari et al., 2020), microtubule nucleation (Woodruff et al., 2017), and the adaptive stress response (Franzmann & Alberti, 2019).

The nucleolus is a large and complex subnuclear compartment that can also be considered a multicomponent and multilayered biomolecular condensate (Lafontaine et al., 2021). It is formed around arrays of ribosomal gene repeats (rDNA), which are transcribed by RNA polymerase I (RNA Pol I) to produce ribosomal RNA (rRNA) (Cerqueira & Lemos, 2019). In the nucleolus, RNAs and hundreds of different proteins are segregated into three immiscible phases: the fibrillar center (FC), which is the RNA Pol I transcription factory; the dense fibrillar component (DFC), which is the intermediate layer rich in fibrillarin (FBL); and the granular component (GC), which is the outer layer rich in nucleophosmin (NPM1/B23). The mechanisms underlying DFC and GC formation through phase separation are relatively well understood. Studies have demonstrated that a ternary mixture comprising the GC protein NPM1, a monomeric version of the DFC protein FBL, and generic rRNA is sufficient to form an FBL-enriched inner phase that coexists with an NPM-enriched outer phase (Feric et al., 2016; Mitrea et al., 2016; Mitrea et al., 2018). In contrast, the mechanisms of FC formation and the role of LLPS in this process have received considerably less attention.

Previous studies on FC organization in living cells suggest that FC assembly may also occur via phase separation (Falahati et al., 2016; Falahati & Wieschaus, 2017). Several lines of evidence support this hypothesis. Firstly, FC components migrate to the nucleolar periphery and aggregate into large structures known as nucleolar caps when RNA Pol I transcription is inhibited by actinomycin D (AMD) or genotoxic drugs (Harding et al., 2015; Korsholm et al., 2019; Reynolds et al., 1964). This structural transformation is reminiscent of liquid droplet fusion. Secondly, high-speed molecular dynamics indicative of fluid-like behavior have been observed in nucleolar caps using single-molecule tracking of RNA Pol I and chromatin-bound upstream binding transcription factor (UBFT/UBF) (Ide et al., 2020). Thirdly, both UBF and its partner treacle ribosome biogenesis factor 1 (TCOF1/Treacle) demonstrated the ability to undergo condensation in vitro and in vivo, respectively (Jaberi-Lashkari et al., 2023; King et al., 2024). Therefore, it is likely that FC assembly and organization rely on LLPS, similar to other nucleolar subcompartments. However, further studies are needed to elucidate the molecular determinants and biophysical properties underlying this process.

In this context, the nucleolar phosphoprotein Treacle is of particular interest. As previously mentioned, it directly interacts with UBF and RNA Pol I, colocalizing with them within the FC. In addition, Treacle exhibits both transcription-dependent and transcription-independent functions (Gal et al., 2022). Its role in ribosome biogenesis is well-documented since it assists in the transcription and processing of rRNA (Gonzales et al., 2005; Lin & Yeh, 2009; Valdez et al., 2004). It is also a master regulator of the DNA damage response (DDR) in the nucleolus, mediating the recruitment of repair factors to rDNA damaged by I-PpoI-induced breaks, replication stress, or R-loop formation (Ciccia et al., 2014; Korsholm et al., 2019; Larsen et al., 2014; Mooser et al., 2020; Velichko et al., 2019, 2021). Therefore, it acts as a nucleolar hub with both transcriptional and DNA repair functions.

The multifunctionality of Treacle may be related to its ability to engage in multivalent interactions and undergo phase separation because of the presence of extended intrinsically disordered regions. In addition, Treacle’s partners in transcription (RNA Pol I, UBF) and DNA repair (DNA topoisomerase II binding protein 1 [TOPBP1]) also utilize phase separation as a functioning mechanism (Frattini et al., 2021; Ide et al., 2020; King et al., 2024). Transcription of rDNA can be regulated by phase separation of RNA Pol I, while TOPBP1 condensation can activate the ATR serine/threonine kinase (ATR) signaling pathway (Frattini et al., 2021; Ide et al., 2020).

Considering these observations, we conducted an in-depth study of Treacle’s condensation characteristics and their impact on the structural organization and functional dynamics of nucleoli. We found that Treacle can form biomolecular condensates and characterized its structural features regulating this process. Through its ability to condense, Treacle promotes the formation of nucleolar FCs by recruiting and concentrating transcription factors at rDNA and separating FC components from the DFC. Collectively, this segregates rRNA synthesis and subsequent processing in distinct compartments of the nucleolus. Impairing Treacle’s condensation ability results in the mixing of FC and DFC components, reducing the efficiency of both processes, equivalent to completed depleting endogenous Treacle. We also demonstrate that Treacle’s phase separation is critical for its interaction with TOPBP1 and activation of the DDR in rDNA during genotoxic stress. Our findings reveal the role of Treacle not only as a structural scaffold for FCs but also as a nucleolar hub integrating the functions of rDNA transcription, rRNA processing, and preserving rDNA integrity.

Results

Treacle is a scaffold protein for nucleolar FCs and DFCs

First, we defined the role of Treacle in the structural organization of the nucleolus. We employed the CRISPR/Cas9 system to knock out the TCOF1 gene encoding Treacle in HeLa cells to obtain a population of cells lacking Treacle signals (Treacle-negative cells; Fig. S1A). Immunocytochemical staining for UBF and RNA Pol I subunit A (POLR1A/RPA194; FC markers), FBL (DFC marker), and B23 and nucleolin (NCL; GC markers) revealed that Treacle-positive cells exhibited a classical tripartite nucleolar structure, with Treacle colocalizing with UBF1 and RPA194 in FCs, FBL in DFCs, and B23 and NCL surrounding DFCs (Fig. 1A). Treacle depletion resulted in the disappearance of distinct FC and DFC structures, causing RPA194, UBF, and FBL to diffuse throughout the nucleolus (Figs. 1B, S1A). This suggests a mixing of FC and DFC components (Fig. S1B). The organization of the GC component, however, was minimally affected (Figs. 1B, S1C). Notably, Treacle depletion also partially redistributed RPA194, UBF, FBL, and NCL from the nucleolus to the nucleoplasm, likely reflecting disrupted ribosomal gene transcription (Figs. 1B, S1A). Similar effects were observed in normal human skin fibroblasts following Treacle depletion (Fig. S1D–E). These results indicated that Treacle plays a critical role in forming the integrated structure of FCs and DFCs.

Treacle is a scaffold protein for nucleolar FCs and DFCs.

(A) Intact HeLa cells (Treacle-positive) were fixed and co-immunostained with Treacle and with either RPA194, UBF, Fibrillarin, B23 or Nucleolin antibodies. DNA was stained with DAPI (gray). Cells were analyzed by laser scanning confocal microscopy. Representative images of cells and nucleoli (magnified images) are shown. Co-localization analysis was performed on the merged images. Graphs illustrate quantification in arbitrary units of fluorescence distribution along the lines shown in the figures. (B) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene. After 7-10 days after transfection, the cells were fixed and co-immunostained with Treacle and either RPA194, UBF, Fibrillarin, B23 or Nucleolin antibodies. DNA was stained with DAPI (gray). Cells were analyzed by laser scanning confocal microscopy. Representative images of Treacle-negative cells and nucleoli (magnified images) are shown. (C) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene (experiment). After 7-10 days after transfection cell were fixed, immunostained with Treacle antibodies and subjected to cell sorting in the fluorescent analysis mode to obtain Treacle-negative populations. Intact HeLa cells (control) were fixed, immunostained with IgG antibodies, passed through all FACS-related procedure in the light scattering analysis mode and used as a control (Treacle-positive). The sorted cell fractions were used for chromatin immunoprecipitation (ChIP) analysis with either Treacle, RPA194 or UBF antibodies. ChIP was followed by qPCR using the d0, d1, d3, d5, d7 primers to the rRNA gene (positioned as indicated on the scheme). Data are represented relative to the input. Values are means ±SD from at least three independent replicates (**, p < 0.01; *, p < 0.1; by unpaired t test). (D) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene. After 7-10 days after transfection the cells were pulsed with EU (100 μM for 2 hr), fixed and immunostained with Treacle antibodies. EU (green) was revealed by click chemistry. The DNA was stained with DAPI (gray). Representative images of Treacle-positive and Treacle-negative HeLa cells are shown. Quantification of EU fluorescence intensities in Treacle-positive and Treacle-negative HeLa cells are show in the right panel (**, p < 0.01 by unpaired t test; n>500). (E) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene (experiment). After 7-10 days after transfection the cells were fixed, immunostained with Treacle antibodies and subjected to cell sorting in the fluorescent analysis mode to obtain Treacle-negative populations. Intact HeLa cells (control) were fixed, stained with IgG antibodies, passed through all FACS-related procedure in the light scattering analysis mode and used as a control (Treacle-positive). The sorted cell fractions were used for RNA extraction. RT-qPCR was performed; it shows levels of 47S pre-rRNA normalized to GAPDH mRNA. Normalized pre-rRNA level in Treacle-positive cells is set to 1. Values are mean ± SD. The calculation is presented for 5 biological replicates (**, p < 0.01 by unpaired t test). (F) Hela cells were processed as described in (E). It shows levels of A’-site contained unprocessed rRNA normalized to GAPDH mRNA. Normalized unprocessed rRNA level in Treacle-positive cells is set to 1. Values are mean ± SD. The calculation is presented for 5 biological replicates (**, p < 0.01 by unpaired t test).

Next, we investigated the impact of nucleolar FC disintegration on the efficiency of rDNA transcription and the recruitment of the RNA Pol I transcription machinery to rDNA. Chromatin immunoprecipitation analysis (ChIP) of cells transfected with an anti-Treacle single guide RNA (sgRNA) revealed a decreased occupancy of UBF and RPA194 within the rDNA coding region (Fig. 1C). In contrast, transfection with a mock sgRNA did not cause such an effect (Fig. S2A). These observations suggest that Treacle amplifies the recruitment rather than initiates the primary binding of the transcription machinery to rDNA. Depleting Treacle substantially suppressed rDNA transcription, although not completely, unlike the repression induced by AMD (Figs. 1D–E, S2B). Furthermore, cells depleted of Treacle exhibited impaired 5′ external transcribed spacer processing, particularly in the cleavage of the A′ site (Fig. 1F), which is one of the initial steps in 18S rRNA maturation.

In summary, Treacle functions as a scaffold protein in nucleolar FCs, and its depletion compromises both rRNA gene transcription and processing.

Treacle drives the formation of biomolecular condensates

Currently, it has become evident that various scaffold proteins can self-organize by forming liquid-phase condensates (Banani et al., 2017; Shin & Brangwynne, 2017). According to protein structure predictors (e.g., AlphaFold, IUPred2, PONDR, and FuzDrop), Treacle is a fully intrinsically disordered protein (Fig. S3A). Nevertheless, Treacle can be broadly categorized into three distinct functional domains: the N-terminal domain (ND), the central domain, and the C-terminal domain (CD) (Gal et al., 2022). The ND of Treacle harbors a LisH motif and a nuclear localization signal (NLS). The central domain contains 15 low-complexity regions (LCRs) enriched in serine and glutamic acid residues (S/E-rich LCRs), which are interspersed with unstructured linker sequences (Jaberi-Lashkari et al., 2023). The 11 linkers within the central domain and their adjacent S/E-rich LCRs exhibit high homology. Consequently, the central domain will be referred to as the repeating domain (RD) henceforth. The CD harbors additional NLSs, a UBF binding site, and a lysine-rich LCR (K-rich LCR) at the terminus, which also functions as a nucleolar localization signal (NoLS) (Fig. 2A).

Treacle drives the formation of nuclear condensates.

(A) The structure of the most common isoform of Treacle. Treacle isoform d (1488 amino acids, 152 kDa, NP_001128715.1) is encoded by the Treacle transcript variant 4. It is an intrinsically disordered protein with N-terminal (ND, 1–83 aa), C-terminal (CD, 1121–1488 aa) regions, and central repeated domain (RD, 83–1121 aa) consisting of 15 low complexity regions (LCR) interspersed with disordered linker sequences. (B) The purified recombinant fragment of Treacle undergoes condensation in vitro in the presence of 5% polyethylene glycol. (C) Condensates of recombinant Treacle’s fragment were time-lapse imaged. Representative images of condensates fusion events are shown. (D) HeLa cells were transfected with Treacle-FusionRed-Cry2 (opto-Treacle) construct. DNA was stained with Hoechst 33342 (gray). Formation of Treacle condensates was induced with blue light illumination of the cells for 10 sec. Representative images are shown. (E) HeLa cells were transfected with Treacle-Katushka2S (Treacle-2S) at a quantity of 50 ng plasmids per 2×105 cells. For low levels of expression analysis, the cells were fixed 16-24 hours after transfection. For high levels of expression analysis, the cells were fixed 48 hours after transfection. DNA was stained with DAPI (gray). Cells were analyzed by laser scanning confocal microscopy. Surface reconstructions of Treacle-2S FCs or extranucleolar condensates are shown in the right panels. Graphs illustrate quantification in arbitrary units of the 3D analysis of Treacle-2S condensate’s shape by LasX software. Sphere form factor = sphere area/particle area. (F) 16-24 h after transfection with Treacle-2S, HeLa cells were treated with 0.05µg/ml actinomycin D (AMD) to induce rDNA transcriptional repression and time-lapse imaged for 120 minutes. DNA was stained with Hoechst 33342 (gray). (G) 48 h after transfection with Treacle-2S, HeLa cells were live time-lapse imaged for 12 hours. DNA was stained with Hoechst 33342 (gray). Representative images of cells with Treacle-2S condensates fusion events are shown. Graphs illustrate the frequency of fusion Treacle condensates per cell for 12 hours. The calculation is presented for 6 biological replicates of 10 cells each. (H) HeLa cells were transiently transfected with Treacle-2S. 16-24 h after transfection half-FRAP analysis of Treacle-labelled fibrillar center (FCs) was performed. Half of the one FC was bleached, and fluorescence recovery was monitored. Representative time-lapse images of the photobleached FC are shown (magnified images). DNA was stained with Hoechst 33342 (gray). (I) Graphs illustrate the quantification of the FCs half-FRAP analysis described in (H). Half-FRAP curves represent the normalized intensity in the bleached half (blue) and the non-bleached half (green). Each trace represents an average of measurements for at least twenty FCs; error bars represent SD. (J) 16-24 h after transfection with Treacle-2S, HeLa cells were treated with 0.05µg/ml AMD to induce the formation of nucleolar caps. Half-FRAP analysis of Treacle-labelled nucleolar caps was performed. Representative time-lapse images of photobleached nucleolar caps are shown (magnified images). DNA was stained with Hoechst 33342 (gray). (K) Graphs illustrate the quantification of the nucleolar caps half-FRAP analysis described in (J). Half-FRAP curves represent the normalized intensity in the bleached half (blue) and the non-bleached half (green). Each trace represents an average of measurements for at least twenty caps; error bars represent SD. (L) HeLa cells were transiently transfected with Treacle-2S. 48 h after transfection, half-FRAP analysis of Treacle-2S extranucleolar condensates was performed. Half of one condensate was bleached, and fluorescence recovery was monitored. The analysis included condensates with diameters of up to 3 µm (type 1), as well as the largest condensates exceeding 5 µm in diameter (type 2). Representative time-lapse images of photobleached Treacle-2S extranucleolar condensates of both categories are shown (magnified images). DNA was stained with Hoechst 33342 (gray). (M) Graphs illustrate the quantification of the nucleolar caps half-FRAP analysis described in (L). Half-FRAP curves represent the normalized intensity in the bleached half (blue) and the non-bleached half (green). Each trace represents an average of measurements for at least twenty extranucleolar condensates of both categories; error bars represent SD. DNA was stained with Hoechst 33342 (gray).

To evaluate the ability of Treacle to self-assemble into biomolecular condensates, we first investigated its condensation properties using an in vitro system. For this purpose, a fragment of the Treacle protein (amino acids 291–426), encompassing two S/E-rich LCRs and two linker regions, was purified from Escherichia coli. In vitro experiments demonstrated that this recombinant fragment formed liquid-like condensates in the presence of 5% polyethylene glycol (Fig. 2B). These condensates exhibited a spherical morphology and frequently underwent fusion events (Fig. 2C).

To investigate Treacle’s ability to self-organize into biomolecular condensates within cells, we generated fusion proteins by linking Treacle to FusionRed (a non-self-dimerizing fluorescent protein) and Arabidopsis thaliana cryptochrome 2 (Cry2), which is known to oligomerize upon exposure to blue light, facilitating the formation of “optoDroplets” (Shin & Brangwynne, 2017). Expression of this opto–Treacle chimeric protein in HeLa cells revealed foci formation in nucleoli in the absence of blue light, consistent with Treacle’s natural tendency to occupy FCs (Fig. 2D; see Fig. S3B for the controls). Under blue light, opto–Treacle formed multiple small foci throughout the nucleoplasm (Fig. 2D).

Next, we explored Treacle’s ability to form biomolecular condensates when overexpressed as a fusion with the far-red fluorescent protein Katushka2S (referred to as 2S; Shcherbo et al., 2007) or green fluorescent protein (GFP). At low expression levels (16–24 hours post-transfection), Treacle’s fusion protein formed foci only in nucleoli, reflecting its natural occupation of FCs (Figs. 2E top panel, S3C). However, at increased expression levels (48–72 hours post-transfection), it began to form large spherical structures in the nucleoplasm (Figs. 2E bottom panel, S3D). The formation of these structures could not be attributed to the oligomerization of 2S or GFP, as comparable large nuclear condensates were observed at high levels of Treacle expression, even in the absence of a fused fluorescent protein (Fig. S3E). Both intranucleolar and extranucleolar Treacle–2S foci, induced by low and high levels of fusion protein expression respectively, exhibited a spherical shape with an aspect ratio close to one, suggesting susceptibility to surface tension (Fig. 2E). Transmission electron microscopy confirmed the internal homogeneity of Treacle condensates, showing circular structures with homogeneous protein content without any electron-dense or -light regions inside the condensates (Fig. S3F).

Like the in vitro-formed Treacle condensates, intracellular Treacle condensates demonstrated the ability to fuse. Our observations revealed that nucleolar caps form during AMD-induced (Figs. 2F, S3G; Movie S1) or DNA damage-induced (Fig. S3H; Movie S2) rDNA transcriptional repression due to the fusion of intranucleolar Treacle–2S foci. Prolonged live-cell imaging further demonstrated that extranucleolar biomolecular condensates from highly overexpressed Treacle–2S frequently fuse, indicating dynamic clustering (Fig. 2G; Movie S3).

Liquid condensates exhibit a high molecular exchange rate, often assessed through fluorescence recovery after photobleaching (FRAP; Ganser & Myong, 2020). To analyze the molecular dynamics of Treacle within its condensates, we employed a combination of full- and half-FRAP methods. Half-FRAP is a specific variation of partial FRAP, where half of the structure of interest is photobleached, unlike full-FRAP, which involves photobleaching the entire structure (Muzzopappa et al., 2022). Half-FRAP analysis revealed an increase in signal intensity in the bleached half of Treacle-2S-formed FCs, accompanied by a proportional decrease in fluorescence in the non-bleached half (Fig. 2H, I; Movie S4). Full-FRAP analysis of Treacle dynamics within FCs revealed a slow recovery rate (Fig. S4A), suggesting preferential internal mixing. This indicates that molecular exchange occurs between the two halves of the condensate without significant exchange across the boundary separating the condensate from the surrounding phase. Notably, fluorescence in the non-bleached region decreased to approximately half its initial value, suggesting near-complete internal mixing within the coacervate. Comparable full- and half-FRAP dynamics were observed for AMD-induced nucleolar caps (Figs. 2J, K, S4B; Movie S5) and extranucleolar condensates formed under elevated Treacle–2S expression (Figs. 2L, M, S4C; Movie S6). Furthermore, the largest extranucleolar condensates exhibited reduced Treacle molecular dynamics, suggesting more gel-like properties (Figs. 2L, M, S4D; Movie S7). Presumably, Treacle condensates undergo a liquid-to-gel phase transition over time or upon reaching a critical protein concentration within the condensate. Therefore, using diverse model systems, we illustrated that Treacle can generate biomolecular condensates with key features of structures formed through liquid-like phase separation.

Treacle’s phase separation is regulated by its central and C-terminal domains

Since we have shown that an RD containing two S/E-rich LCRs can form liquid-like condensates in vitro, we next sought to investigate to what extent this condensation ability of RD is exerted intracellularly and to determine the roles of the CD and ND in intracellular Treacle condensation. For this purpose, we generated Treacle mutants missing the ND (Δ1–83), RD (Δ83–1121), or CD (Δ1121–1488) and overexpressed them in HeLa cells. We then assessed the condensation ability of the overexpressed deletion mutants in wild-type cells using several intracellular condensate models: the formation of (i) FCs and (ii) nucleolar caps under low expression levels (24 hours post-transfection) and (iii) large extranucleolar condensates under high expression levels (48 hours post-transfection).

The Δ1–83 mutant demonstrated condensation properties indistinguishable from full-length Treacle: it concentrated in FCs, formed stable nucleolar caps during transcriptional repression, and, upon increased expression levels, generated numerous large extranucleolar condensates (Figs. 3A, S5A, D, E). Using partial FRAP analysis, we further demonstrated that, similar to full-length Treacle (Fig. S5B, C), the Δ1–83 mutant exhibited a high rate of molecular exchange in all types of model condensates (Fig. 3B, C).

Treacle’s condensation is regulated by its central and C-terminal domains

(A) HeLa cells were transfected with Treacle-2S Δ1-83 deletion mutant. For low or high levels expression analysis cells were cultivated 16-24 or 48 h after transfection respectively. The expression level is indicated by the colored zone to the left of the cell images. HeLa cells with low expression level were additionally treated with CX-5461 to induce rDNA transcriptional repression and subsequent nucleolar cap formation. Cells were fixed and analyzed by laser scanning confocal microscopy. DNA was stained with DAPI (gray). Representative images of Treacle-2S Δ1-83 condensate are shown. (B) HeLa cells were transfected with Treacle-2S Δ1-83 deletion mutant and processed as described in (A). Partial FRAP analysis of Treacle-2S Δ1-83 condensates was performed. A part of each condensate type was photobleached, and the subsequent fluorescence recovery was monitored. Representative time-lapse images of the photobleached condensates are shown (magnified images). DNA was stained with Hoechst 33342 (gray). (C) HeLa cells were transfected with Treacle-2S Δ1-83 deletion mutant and processed as described in (A). Graphs illustrate the quantification of the Treacle-2S Δ1-83 condensates partial FRAP analysis described in (B). Each trace represents an average of measurements for at least twenty Treacle Δ1-83 condensates of each type; error bars represent SD. (D) HeLa cells were transfected with Treacle-2S Δ83-1121 deletion mutant and processed as described in (A). (E) HeLa cells were transfected with Treacle-2S Δ83-1121 deletion mutant and processed as described in (A). Partial FRAP analysis of Treacle-2S Δ83-1121 condensates dynamics was performed as described in (B). (F) HeLa cells were transfected with Treacle-2S Δ83-1121 deletion mutant and processed as described in (A). Partial FRAP analysis of 2S-fused Treacle Δ83-1121 condensates dynamics was performed as described in (B). Graphs illustrate the quantification of the Treacle-2S Δ83-1121 condensates partial FRAP analysis. Each trace represents an average of measurements for at least twenty Treacle-2S Δ83-1121 condensates of each type; error bars represent SD. (G) HeLa cells were transfected with Treacle-2S Δ1121-1488 deletion mutant fused with nuclear localization signal (NLS) from simian virus 40 (SV40). For low or high levels expression analysis cells were cultivated 16-24 or 48 h after transfection respectively. The expression level is indicated by the colored zone to the left of the cell images. Cells were fixed and analyzed by laser scanning confocal microscopy. DNA was stained with DAPI (gray). Representative images of cells are shown.

As expected, the Δ83–1121 mutant exhibited significantly compromised condensation properties. Under low expression levels, it failed to concentrate in FCs (Figs. 3D, S5F) and diffusely localized in the nucleolus while still maintaining a high molecular dynamics rate (Fig. 3E, F). During rDNA transcriptional repression, the Δ83–1121 mutant did not form classical nucleolar caps but partially redistributed between the nucleolar periphery and nucleoplasm, where it began to form condensates (Figs. 3D, S5F). Surprisingly, such relocation was associated with a change in its molecular dynamics state. Partial FRAP analysis revealed that nucleoplasmic condensates of the Δ83–1121 mutant, formed due to nucleolar transcriptional repression, began to transition into a solid-like state (Fig. 3E, F). A similar liquid-solid transition was observed for the Δ83–1121 mutant under high expression levels. In this case, it no longer formed distinct extranucleolar condensates but merged into a unified pan-nuclear solid network (Figs. 3D–F, S5F). This behavior suggests that the RD deletion alters Treacle’s interaction patterns, making its phase separation properties more reliant on rRNA and other nucleolar proteins.

Unexpectedly, the Δ1121–1488 mutant also failed to condense. When equipped with an artificial NLS from the SV40 virus, it successfully translocated into the nucleus; however, it remained diffusely distributed throughout the nucleoplasm, even under high expression conditions (Figs. 3G, S5G). Therefore, without the CD, Treacle cannot undergo autonomous condensation, likely relying on the CD as a nucleation factor. Collectively, these results suggest that the intracellular Treacle condensation depends on a cooperative interaction between RD and CD.

To gain further insights into these phenomena, we investigated the nature of interactions driving Treacle protein condensation. We assessed the behavior of Treacle intracellular condensates after treatment with 1,6-hexanediol, which disrupts hydrophobic interactions, or high salt concentrations, which interfere with electrostatic interactions. Treatment with the cell-permeable salt ammonium acetate disrupted both FCs and extranucleolar condensates formed at low and high Treacle expression levels, respectively (Fig. 4A). In contrast, treatment with 1,6-hexanediol did not affect the integrity of Treacle condensates (Fig. 4A). These observations were further validated using a condensation model based on the recombinant fragment of Treacle’s RD in vitro. The efficiency of Treacle’s fragment condensation in vitro remained unchanged in the presence of 1,6-hexanediol, whereas increasing salt concentration prevented condensation (Fig. 4B). Additionally, condensation of the Treacle RD fragment in vitro was sensitive to changes in pH. The optimal efficiency of condensation was observed at pH 7.0, whereas acidification of the medium decreased condensation, and alkalization resulted in the generation of insoluble aggregates (Fig. 4B). Therefore, it is clear that Treacle condensation, both in vitro and in vivo, is primarily governed by electrostatic interactions rather than hydrophobic forces.

The condensation of Treacle is controlled by the specific charge distribution in its domains.

(A) HeLa cells were transfected with Treacle-2S. For low or high levels expression analysis, cells were cultivated 16-24 or 48 h after transfection respectively. The expression level is indicated by the colored zone to the left of the cell images. Cells were treated with 5% 1,6-hexandiol (1,6-HEX) for 10 min or 200 mM ammonium acetate for 5 min. Cells were fixed and analyzed by laser scanning confocal microscopy. DNA was stained with DAPI (gray). Representative images of cells are shown. (B) The purified recombinant fragment of Treacle undergoes condensation in vitro in the presence of 10% 1,6-hexandiol (1.6-HEX), 500 mM sodium chloride or buffers with different pH. In each case, 5% PEG 8000 was used as a crowding agent. (C) HeLa cells were transfected with Treacle-2S ΔSE mutant. For low or high levels expression analysis cells were cultivated 16-24 or 48 h after transfection respectively. The expression level is indicated by the colored zone to the left of the cell images. HeLa cells with low expression level were additionally treated with CX-5461 to induce rDNA transcriptional repression. Cells were fixed and analyzed by laser scanning confocal microscopy. DNA was stained with DAPI (gray). Representative images of Treacle ΔSE condensate are shown. (D) HeLa cells were transfected with Treacle-2S ΔSE mutant and processed as described in (C). Partial FRAP analysis of Treacle-2S ΔSE condensates was performed. A part of each condensate type was photobleached, and the subsequent fluorescence recovery was monitored. Representative time-lapse images of the photobleached condensates are shown (magnified images). DNA was stained with Hoechst 33342 (gray). (E) HeLa cells were transfected with Treacle-2S ΔSE mutant and processed as described in (C). Graphs illustrate the quantification of the Treacle-2S ΔSE condensates partial FRAP analysis described in (D). Each trace represents an average of measurements for at least twenty Treacle-2S ΔSE condensates of each type; error bars represent SD. (F) HeLa cells were transfected with Treacle-2S Δ1350-1488 deletion mutant (Treacle-2S ΔNoLS) and processed as described in (C). (G) HeLa cells were transfected with Treacle-2S ΔNoLS deletion mutant and processed as described in (C). Partial FRAP analysis of Treacle-2S ΔNoLS condensates was performed as described in (D). (H) HeLa cells were transfected with Treacle-2S ΔNoLS deletion mutant and processed as described in (C). Partial FRAP analysis of Treacle-2S ΔNoLS condensates dynamics was performed as described in (D). Graphs illustrate the quantification of the Treacle-2S ΔNoLS condensates partial FRAP analysis. Each trace represents an average of measurements for at least twenty Treacle ΔNoLS condensates of each type; error bars represent SD. (I) Charge plots of full-length Treacle (left panel) and charge-scrambled Treacle (Treacle CS) form (right panel) are shown. Positive and negative charge blocks are depicted by blue and red triangles, respectively. Charge distribution was calculated as the sum of the charges (Arg and Lys, +1; Glu and Asp, −1;) in the 25 amino acids window range. The center of the panel shows the aligned amino acid sequences of one of the S/E-rich LCR and its adjacent linker of RD for both full Treacle and Treacle CS. (J) HeLa cells were transfected with Treacle-2S CS mutant and processed as described in (C).

It is reasonable to hypothesize that electrostatic interactions within Treacle condensates are facilitated by the presence of negatively charged S/E-rich LCRs in the RD and positively charged K-rich LCR in the CD. To evaluate this hypothesis, we generated a Treacle mutant lacking the 13 S/E-rich LCRs in the RD (ΔSE) and a Treacle mutant missing the 1350–1488 region (ΔNoLS), which includes the K-rich LCR. The deletion of the S/E-rich LCRs significantly compromised Treacle’s condensation properties. Like the Δ83–1121 mutant, Treacle ΔSE could not form FCs or nucleolar caps (Figs. 4C, S6A), remained in a liquid state within the nucleolus (Fig. 4D, E), and underwent a liquid-solid phase transition during nucleolar transcriptional repression or high overexpression levels (Fig. 4D, E). These findings indicate that the RD determines Treacle’s condensation properties through the S/E-rich LCRs. In turn, ΔNoLS mutant did not localize to the nucleolus but could form individual perinucleolar condensates (Figs. 4F, S6B). Interestingly, the volume of these condensates increased significantly with the expression level of ΔNoLS, but their number did not (Figs. 4F, S6B). This observation suggests that, in contrast to full-length Treacle, which forms a substantial number of extranucleolar condensates at high expression levels, ΔNoLS condensates exhibit a limited number of nucleation sites. This implies that the positively charged lysines in the K-rich LCR of the CD act as nucleation points for autonomous Treacle condensate formation. It is also noteworthy that all types of ΔNoLS condensates displayed high molecular dynamics (Fig. 4G, H). In contrast to full-length Treacle condensates, molecular exchange within ΔNoLS condensates occurred across the boundary separating the coacervate from the surrounding phase, rather than through internal mixing (Fig. S6C, D). This finding reinforces the role of the K-rich LCR as both a nucleation element in Treacle condensate formation and a regulator of Treacle’s volumetric dynamics within these condensates.

The disruption of Treacle’s condensation properties following the deletion of the S/E-rich LCRs underscores the significance of the negative charge within the central domain for phase separation. However, as noted earlier, the S/E-rich LCRs in the RD are not clustered but rather dispersed throughout the RD, interspersed with linker regions. We propose that for Treacle to exhibit proper condensation behavior, it is not merely the presence of a negative charge in the RD that is critical, both the presence of a negative charge in the RD and its specific spatial distribution within the molecule are critical.. Indeed, an analysis of the charge distribution along the Treacle amino acid sequence demonstrated that each S/E-rich LCR, in conjunction with its adjacent linker, within the RD constitutes a strong diblock ampholyte, where a positively charged block (linker) is followed by a negatively charged block (S/E-rich LCR; Fig. 4I left panel). In order to determine the physiological importance of this charge distribution, a variant of Treacle (Treacle CS) was created with the same overall net charge but scrambled blocks (Fig. 4I right panel). In Treacle CS, regions of opposite charge were removed while maintaining the same overall isoelectric point, amino acid composition, and positions of all other residues. As expected, scrambling the charges in the RD fully reproduced the condensation-defective phenotype observed in the Δ83–1121 and ΔSE mutants. Treacle CS failed to form FCs or nucleolar caps, instead assembling into a single pan-nuclear network at high overexpression levels (Fig. 4J, S6E). These findings suggest that Treacle’s condensation properties depend not only on the net negative charge within the RD but are critically influenced by the specific distribution of charges within this region.

In summary, our findings demonstrate that the cooperative interaction between the RD and CD governs the phase separation properties of Treacle, whereas the ND plays a negligible role in this process. The positively charged K-rich LCR within the CD facilitates its role as a nucleation site for Treacle condensate formation, while the alternating charge blocks in the RD stabilize Treacle condensates, ensuring that Treacle behavior remains independent of subnuclear localization or nucleolar transcriptional activity.

Treacle condensation is essential for its proper interaction with nucleolar subcompartments

We next aimed to investigate the role of Treacle’s condensation properties in its functional interactions with the components of the FC, DFC, and GC within the nucleolus. First, we significantly downregulated endogenous Treacle expression (Treacle kd) using RNA interference (Fig. S7A) and then analyzed the structure organization of the FC, DFC, and GC within the nucleolus in the context of small interfering RNA (siRNA)-resistant full-length or condensation-defective (Δ83–1121 or CS) Treacle mutants.

The expressed full-length Treacle exhibited normal nucleolar localization, similar to endogenous Treacle. It co-localized within the FC with RPA194 and UBF, while FBL was localized in the DFC at the periphery of the FC, and NCL and B23 surrounded the DFC (Fig. 5A). In contrast, expression of the Δ83–1121 or CS Treacle mutants led to the dispersal and mixing of FC and DFC components within the nucleolus (Figs. 5B, S7B). Despite this, the Δ83–1121 or CS mutants still co-localized with UBF1, but to a significantly lesser extent with RPA194, suggesting a potential loss of their binding interaction. Interestingly, these mutants showed partial overlap with NCL, but not with B23 (Figs. 5B, S7B). These observations suggest that disrupting Treacle’s condensation properties may alter its interaction preferences..

The condensation of Treacle facilitates its functional interactions with partner proteins

(A) Endogenous Treacle was depleted by siRNA-mediated knockdown (Treacle kd). Next, Treacle-depleted cells were transfected with siRNA-resistant full-length Treacle-2S (Treacle-2S full). Cells were fixed 16-24 after transfection and immunostained with either RPA194, UBF, Fibrillarin, B23 or Nucleolin antibodies. Cells were analyzed by laser scanning confocal microscopy. Representative images of magnified nucleoli are shown. Co-localization analysis was performed on the merged images. Graphs illustrate quantification in arbitrary units of Treacle-2S variants and RPA194, UBF, Fibrillarin, B23 or Nucleolin fluorescence distribution along the lines shown in the figures. (B) Endogenous Treacle was depleted by siRNA-mediated knockdown (Treacle kd). Next, Treacle-depleted cells were transfected with siRNA-resistant charge-scrambled Treacle-2S (Treacle-2S CS) and processed as described in (A). (C) Endogenous Treacle was depleted by siRNA-mediated knockdown (Treacle kd). Next, Treacle-depleted cells were transfected with siRNA-resistant full-length Treacle-2S (Treacle-2S full). 16-24 h after transfection cells were treated with 0.05µg/ml actinomycin D (AMD) to induce rDNA transcriptional repression, fixed and processed as described in (A). (D) Endogenous Treacle was depleted by siRNA-mediated knockdown (Treacle kd). Next, Treacle-depleted cells were transfected with siRNA-resistant charge-scrambled Treacle-2S (Treacle-2S CS). 16-24 h after transfection cells were treated with 0.05µg/ml actinomycin D (AMD) to induce rDNA transcriptional repression, fixed and processed as described in (A).

This hypothesis is further supported by the analysis of the localization of nucleolar components and condensation-defective Treacle forms following AMD-induced nucleolar transcriptional repression. After AMD treatment, full-length Treacle co-localized with UBF1 and RPA194 at the nucleolar caps (Fig. 5C). FBL partially merged with the caps, while NCL and B23 were completely displaced to the nucleolar rim and nucleoplasm (Fig. 5C). In contrast, the AMD-induced perinucleolar or nucleoplasmic solid condensates formed by the Δ83–1121 or CS Treacle mutants still co-localized with UBF1, but to a significantly lesser extent with RPA194 (Figs. 5D, S7C). Interestingly, these condensates exhibited behavior opposite to that of full-length Treacle condensates with respect to FBL and NCL: they displaced FBL and incorporated NCL (Figs. 5D, S7C). The patterns of association between FC/DFC/GC proteins and the extranucleolar condensates of Δ83–1121 and CS Treacle mutants at high expression levels exhibited similar alterations when compared to those observed with full-length Treacle (Fig. S7D, E).

Thus, the disruption of Treacle’s condensation properties, either through RD deletion or charge scrambling, results in a shift in its interaction repertoire. Specifically, while interaction with UBF1 is preserved, interactions with RNA Pol I and FBL are notably diminished, whereas a new interaction with NCL is established. It is likely that the correct condensation of Treacle contributes to the formation of the cooperative FC/DFC structure by mediating functional interactions with partner proteins.

The condensation of Treacle regulates rRNA transcription and processing

In the above sections, we demonstrated that Treacle supports both rRNA transcription and processing. We hypothesized that this multifunctionality of Treacle could be related to its ability to condense and the formation of the cooperative FC/DFC structure. To test this hypothesis, we significantly downregulated endogenous Treacle expression (Treacle kd) using RNA interference and then analyzed rRNA transcription levels in the context of siRNA-resistant full-length or condensation-defective (Δ83– 1121 or CS) Treacle mutants. We found that the expression of condensation-defective Treacle mutants significantly reduced the transcription of rRNA compared to full-length Treacle (Fig. 6A). This effect correlated with a reduced ability of condensation-defective Treacle mutants to maintain the association of partner proteins RPA194 and UBF1 at the rDNA promoter (Fig. 6C), along with diminished intrinsic binding of these mutants to rDNA (Fig. 6B).

Treacle LLPS is essential for the transcription and processing of rRNA

(A) Endogenous Treacle was depleted by siRNA-mediated knockdown (Treacle kd). Next, Treacle-depleted cells were transfected with siRNA-resistant full-length Treacle-2S (Treacle-2S full), charge-scrambled Treacle-2S (Treacle-2S CS) or Treacle-2S Δ83-1121 deletion mutant (Treacle-2S Δ83-1121). Cells were fixed 16-24 after transfection and subjected to cell sorting in the fluorescent analysis mode to obtain 2S-positive populations. The sorted cell fractions were used for RNA extraction. RT-qPCR was performed; it shows levels of 47S pre-rRNA normalized to GAPDH mRNA. Normalized pre-rRNA level in full-length Treacle-2S-positive cells is set to 1. Values are mean ± SD. The calculation is presented for 5 biological replicates. **, p < 0.01 by unpaired t test. (B) HeLa cells were processed as described in (A). Cells were fixed 16-24 after transfection and subjected to ChIP-seq analysis with Katushka2S antibodies. ChIP-seq signal were normalized to the input. (C) HeLa cells were processed as described in (A). Cells were fixed 16-24 after transfection and subjected to cell sorting in the fluorescent analysis mode to obtain 2S-positive populations. The sorted cell fractions were used for used for ChIP with RPA194 or UBF antibodies. ChIP was followed by qPCR using the d1 primers to the promoter of the rRNA gene (positioned as indicated on the scheme). Percentage of enrichment relative to input in full-length Treacle-2S-positive cells is set to 1.Values are mean ± SD. The calculation is presented for 5 biological replicates. **, p < 0.01 by unpaired t test. (D) HeLa cells were processed as described in (A). Cells were fixed 16-24 after transfection and stained for 47S pre-rRNA (revealed by single-molecule FISH, smFISH). Cells were analyzed by laser scanning confocal microscopy. Representative images of magnified nucleoli are shown. Co-localization analysis was performed on the merged images. Graphs illustrate quantification in arbitrary units of Treacle-2S variants and smFISH fluorescence distribution along the lines shown in the figures. (E) HeLa cells were transfected with full-length Treacle-2S. 48 h after transfection cells were fixed and stained for 47S pre-rRNA (revealed by single-molecule FISH, smFISH). Cells were analyzed by laser scanning confocal microscopy. Representative images of magnified nucleoli are shown. (F) HeLa cells were transfected with a construct coding CRISPR/Cas9 and sgRNA to the TCOF1 gene (Treacle kn). After 7-10 days after transfection, the cells were fixed and stained with UBF antibodies and 47S pre-rRNA. Representative images of Treacle-negative nucleolus (magnified images) are shown. (G) HeLa cells were processed as described in (A). Cells were fixed 16-24 after transfection and subjected to cell sorting in the fluorescent analysis mode to obtain 2S-positive populations. The sorted cell fractions were used for RNA extraction. RT-qPCR was performed; it shows levels of A’-site contained unprocessed rRNA normalized to GAPDH mRNA. Normalized unprocessed rRNA level in full-length Treacle-2S-positive cells is set to 1. Values are mean ± SD. The calculation is presented for 5 biological replicates. **, p < 0.01 by unpaired t test.

These observations suggest that the condensation properties of Treacle underlie its ability to concentrate components of the transcriptional machinery in the FC and separate them from the DFC. Moreover, such separation is likely required to maintain highly efficient transcription of rDNA. We hypothesized that the same principle could regulate the role of Treacle in rRNA processing. To test this assumption, we depleted the level of endogenous Treacle using RNA interference and then analyzed the spatial localization of newly synthesized 47S pre-rRNA and its processing level in cells expressing siRNA-resistant full-length or condensation-defective (Δ83–1121 or CS) Treacle mutants. Microscopy analyses with fluorescence in situ hybridization (FISH)-labeled rRNA revealed that newly synthesized 47S pre-rRNA transcripts clustered at the periphery of FCs in full-length Treacle-expressing cells, reflecting the radial flow of rRNA from the FC to the GC (Fig. 6D). Newly synthesized rRNA was also relocated to the periphery of large Treacle condensates formed at high expression levels (Fig. 6E). However, no apparent clustering of newly synthesized rRNA was observed in cells expressing condensation-defective Treacle mutants. Instead, the newly synthesized transcripts were diffusely mixed with Δ83–1121 and CS Treacle (Fig. 6D). Nascent rRNA transcripts were similarly mixed with delocalized UBF in cells depleted of endogenous Treacle (Treacle kn; Fig. 6F). Finally, reverse transcription-quantitative polymerase chain reaction (RT-qPCR) analysis confirmed that cells overexpressing condensation-defective Treacle mutants had impaired 5′ external transcribed spacer processing compared to cells overexpressing full-length Treacle (Fig. 6G). Therefore, it can be concluded that the mixing of FC and DFC components due to the expression of condensation-defective Treacle or the depletion of endogenous Treacle disrupts the directional traffic of nascent rRNA from the FC to the DFC and further to the GC, potentially causing inefficient processing.

Our results indicate that Treacle’s condensation properties separate the FC and DFC, leading to the spatial segregation of rRNA synthesis and subsequent processing. The mixing of FC and DFC components due to the disruption of Treacle’s condensation ability reduces the efficiency of both processes, with equivalent effects to the complete depletion of endogenous Treacle.

The condensation of Treacle is essential for DDR activation in ribosomal genes under genotoxic stress

Previous studies have demonstrated that Treacle is critical for inducing the DDR in ribosomal genes under certain types of stress (Korsholm et al., 2019; Larsen et al., 2014; Mooser et al., 2020; Velichko et al., 2019, 2021). Here, we aimed to investigate the contribution of Treacle’s condensation to rDNA damage response. We induced rDNA damage using the widely used chemotherapeutic drug etoposide (VP16), which acts as a topoisomerase II inhibitor, inducing double-strand breaks (DSBs) (Bax et al., 2019). Treating cells with 90 µM VP16 for 30 minutes resulted in the rapid recruitment of TOPBP1 to nucleoli and its colocalization with Treacle in HeLa (Figs. 7A, S8A) and MCF7 cells (Fig. S8B). Furthermore, proximity ligation assay (PLA) analysis conducted after DNA damage induction provided evidence of a physical association between TOPBP1 and Treacle (Figs. 7B, S8C). Reducing endogenous Treacle levels, either through RNA interference (Treacle kd) or CRISPR/Cas9-mediated depletion (Treacle kn), effectively blocked the relocalization of TOPBP1 to nucleoli (Figs. 7D, S9A–B) and its occupancy of rDNA (Fig. 7C). Interestingly, with Treacle kd, a substantial amount of Treacle still remained in nucleoli (Fig. 7D). However, this amount was insufficient to facilitate its interaction with TOPBP1, likely due to the need for specific stoichiometric ratios for efficient interaction.

Treacle phase separation is essential for DDR activation in ribosomal genes under genotoxic stress conditions

(A) DMSO-treated and VP16-treated (90 μM, 30 min) HeLa cells were co-immunostained for Treacle (Treacle endogenous; magenta) and TOPBP1 (green) and analyzed by laser scanning confocal microscopy. The DNA was stained with DAPI (gray). Co-localization analysis was performed on the merged images (magnified images). Graphs illustrate quantification in arbitrary units of Treacle and TOPBP1 fluorescence distribution along the lines shown in the figures. Percentage of cells containing TOPBP1 (TOPBP1-positive) foci within nucleoli is shown. (B) DMSO-treated and VP16-treated (90 μM, 30 min) HeLa cells were subjected to Proximity Ligation Assay (PLA) with antibodies against TOPBP1 and Treacle. DNA was stained with DAPI (gray). PLA detection of Treacle-TOPBP1 interactions is visible as distinct green fluorescent dots. (C) Mock-treated HeLa cells or cells with siRNA-mediated Treacle knockdown (Treacle kd) were treated with DMSO or VP16 (90 μM, 30 min). ChIP experiments were performed with antibodies against TOPBP1. ChIP was followed by qPCR using the d1 primers to the promoter of the rRNA gene (positioned as indicated on the scheme). Data are represented relative to the input. Values are means ±SD from at least three independent replicates. (D) Intact HeLa cells, siRNA-depleted for Treacle (Treacle kd) cells or CRISPR/Cas9-depleted for Treacle (Treacle kn) cells were treated with 90 μM VP16 for 30 min. Cells were co-immunostained for TOPBP1 (green) and Treacle (magenta) antibodies and analyzed by laser scanning confocal microscopy. The DNA was stained with DAPI (gray). (E) Endogenous Treacle was depleted by siRNA-mediated knockdown (Treacle kd). Next, Treacle-depleted cells were transfected with plasmid constructs encoding either siRNA-resistant full-length Treacle-2S (Treacle-2S full), Treacle-2S Δ83-1121 deletion mutant, or charge-scrambled mutant Treacle-2S (Treacle-2S CS). 24 h after transfection, cells were treated with VP16 (90 μM for 30 min), fixed and stained for TOPBP1 (green) and analyzed by laser scanning confocal microscopy. The DNA was stained with DAPI (gray). (F) Endogenous Treacle was depleted by siRNA-mediated knockdown (Treacle kd). Treacle-depleted cells were transfected with plasmid constructs encoding either siRNA-resistant Treacle-2S full, Treacle-2S Δ83-1121, or Treacle-2S CS. 24 h after transfection, cells were treated with DMSO or VP16 (90 μM for 30 min) and fixed. Cells were subjected to cell sorting in the fluorescent analysis mode to obtain 2S-positive populations. At least 2×106 sorted cells were used for ChIP with TOPBP1 antibodies. ChIP was followed by qPCR using the d1 primers to the promoter of the rRNA gene (positioned as indicated on the scheme). Data are represented relative to the input. Values are means ±SD from at least three independent replicates. (G) Intact HeLa cells and cells siRNA-depleted for either Treacle (Treacle kd) or TOPBP1 (TOPBP1 kd) were treated with DMSO or 90 μM VP16 for 30 min. ChIP experiments were performed with antibodies against phospho-ATR (pATR; Ser1989), phospho-ATM (pATM; Ser1981) or γH2AX antibodies. ChIP was followed as described in (F). (H) HeLa cells were processed as described in (F). At least 2×106 sorted cells were used for ChIP with phospho-ATR (pATR; Ser1989), phospho-ATM (pATM; Ser1981) or γH2AX antibodies. ChIP was followed by qPCR using the d1 primers to the promoter of the rRNA gene (positioned as indicated on the scheme). Data are represented relative to the input. Values are means ±SD from at least three independent replicates.

Next, we investigated whether the interaction between Treacle and TOPBP1 in Treacle kd cells could be restored by overexpressing different Treacle variants. Immunocytochemical analysis and chromatin immunoprecipitation with antibodies against TOPBP1 indicated that expressing (siRNA)-resistant full-length Treacle in Treacle kd cells effectively restored the VP16-induced interaction between Treacle and TOPBP1 (Figs. 7E, S10A), as well as the enrichment of TOPBP1 at the promoters of rDNA (Fig. 7F). However, expressing (siRNA)-resistant condensation-defective forms of Treacle (Δ83–1121 or CS) did not lead to such restoration (Figs. 7E–F, S10A).

The response to DNA damage involves the activation of various signaling kinases and the recruitment of different repair factors to the DNA break site. Chromatin immunoprecipitation followed by qPCR with antibodies against a panel of DNA repair proteins confirmed that VP16 induced DDR at the promoters of rDNA, including the activation of ATM serine/threonine kinase (ATM) and ATR, phosphorylation of the H2A.X variant histone (H2AX), and recruitment of repair factors such as tumor protein p53 binding protein 1 (TP53BP1/53BP1) and BRCA1 DNA repair associated (BRCA1; Figs. 7G, S10C). As expected, nucleolar DDR was fully controlled by the interaction between Treacle and TOPBP1, as the knockdown of either of these proteins dramatically reduced VP16-induced repair signaling at rDNA (Figs. 7G, S10B–C). Finally, it was confirmed that nucleolar DDR could be efficiently restored in Treacle kd cells by overexpressing (siRNA)-resistant full-length Treacle but not its condensation-defective mutants (Δ83–1121 or CS; Figs. 7H, S10D). Therefore, we can conclude that Treacle’s phase separation is crucial in facilitating its binding to TOPBP1 and activating the DDR in response to genotoxic damage to rDNA.

Discussion

The nucleolus is a multicomponent phase condensate formed on the platform of rRNA, which is processed and incorporated into assembling ribosomes simultaneously with migration from FCs to DFCs and further to GCs. The constitutive components of DFCs and GCs are NCL and NPM1, respectively (Feric et al., 2016; Lafontaine et al., 2021). Which protein plays a comparable role in FCs continues to be debated. Our results suggest that the phosphoprotein Treacle serves as the scaffold protein for nucleolar FCs in human cells. This conclusion aligns entirely with the observation that the evolutionary emergence of Treacle correlates with the appearance of the three-partite nucleolus in amniotes (Jaberi-Lashkari et al., 2023). Introducing human Treacle into zebrafish, which lack a competent ortholog for FC formation, induced the reorganization of the nucleolar structure from bipartite to tripartite, accompanied by the formation of FC-like structures.

Our findings demonstrate that the condensation properties of Treacle are governed by the synergistic interplay between its RD and CD. While the RD can form liquid-like condensates alone in vitro, its condensation within the cellular environment requires the positively charged CD, which likely functions as a nucleation platform. Notably, the zebrafish ortholog of Treacle exhibits sequence homology with human Treacle exclusively within the RD (Hill-Terán et al., 2024), lacking a homologous sequence in the CD. This structural difference likely accounts for its inability to undergo condensation and form FC-like structures.

Treacle functions as a block polyampholyte, with its condensation properties governed by the alternating charges of the CD and the strongly positive charge of the C-terminal domain. These properties suggest that its condensation is likely driven by coacervation or self-coacervation mechanisms. Complex coacervation refers to a phase transition process mediated by polyelectrolyte complexation, where oppositely charged macromolecules associate through electrostatic interactions and phase-separate from the surrounding solution to form dense, charge-balanced structures known as coacervates (Choi et al., 2020; Sathyavageeswaran et al., 2024; Sing & Perry, 2020; Pappu et al., 2023). The driving forces for complex coacervation can stem from homotypic or heterotypic interactions between charged polymers (Sanders et al., 2020). Our experimental findings suggest that both mechanisms may be drivers of Treacle’s coacervation.

Homotypic coacervation of Treacle may occur through a two-step process: nucleation is facilitated by electrostatic interactions between the lysine (K)-rich low-complexity region (LCR) in the CD and the glutamic acid (E)-rich LCR in the RD, while fluidization is achieved via dynamic interactions between the E-rich LCR and K-rich linkers, primarily within the RD. Conversely, heterotypic coacervation of Treacle may involve interactions with other polyampholyte proteins enriched in the FCs, many of which are known Treacle partners and possess the ability to facilitate coacervation. For example, it is hypothesized that the CD of Treacle may promote nucleation through interactions with UBF, which contains extended aspartic acid (D)/glutamic acid (E) tracts and demonstrates autonomous condensation properties in the presence of rDNA (Gal et al., 2022; King et al., 2024; Lin & Yeh, 2009). Additionally, Treacle interacts with RNA Pol I, whose subunits (e.g., POLR1F and POLR1G) are characterized by prominent K-rich regions and E-rich tracts (King et al., 2024). These regions are likely to facilitate electrostatic interactions between RNA Pol I and Treacle, potentially modulating its condensation dynamics within the cellular environment. This hypothesis is further supported by our experimental observations showing that charge scrambling or deletion of Treacle’s RD partially disrupts its interaction with RNA Pol I subunit RPA194, reduces its recruitment to rDNA, and impairs its condensation properties. Together, these findings highlight the interplay of homotypic and heterotypic interactions in driving Treacle’s coacervation and its functional role within the nucleolus.

Endogenous Treacle may undergo post-translational modifications. Specifically, the RD of Treacle is known to be heavily phosphorylated and ubiquitinated (Mooser et al., 2020; Werner et al., 2015, 2018). Currently, the role of these modifications is primarily attributed to facilitating interactions with partner proteins such as the cullin 3 (CUL3), kelch repeat and BTB domain containing 8 (KBTBD8) complex or nucleolar and coiled-body phosphoprotein 1 (NOLC1; Werner et al., 2018). However, it is known that phosphorylation can modulate phase separation, particularly by enhancing or reducing the polyampholytic properties of disordered regions (Yamazaki et al., 2022). We propose that post-translational modifications, such as ubiquitination and phosphorylation, fine-tune the condensation properties of Treacle. These modifications may do so by modulating interactions with binding partners and altering the ampholytic properties of the RD through changes in its charge ratio.

The 47S pre-rRNA is synthesized at the FC/DFC boundary, while its primary processing begins in the DFC. DFC formation is mediated by the phase separation of FBL gathered on the rRNA scaffold and is maintained as long as there is a directed radial flow of rRNA from the FC to the DFC (Wu et al., 2021; Yao et al., 2019). While the role of Treacle in ensuring rRNA processing has been repeatedly described (Calo et al., 2018; Gonzales et al., 2005), the underlying mechanism remains unclear. We demonstrated that the efficiency of rRNA processing is compromised not only by the complete depletion of endogenous Treacle but also by the expression of condensation-defective Treacle mutants. This impairment correlates with a loss of Treacle association with fibrillarin and, conversely, an increased association of Treacle with newly synthesized rRNA and nucleolin. This shift likely disrupts the cooperative organization of the FC/DFC and the mixing of their components. The resulting disorganization and mixing of FC and DFC components, driven by the loss of Treacle’s condensation properties, may compromise the radial flow of rRNA, thereby hindering its normal processing.

We also explored the role of Treacle’s phase separation in nucleolar DDR. Treacle’s well-known partner, TOPBP1, plays a crucial role in activating ATR during DNA replication stress (Day et al., 2021). The interaction between TOPBP1 and Treacle is essential for nucleoli-specific DDR and replication stress response (Mooser et al., 2020; Velichko et al., 2019, 2021). This interaction occurs via phosphoserines 1227 and 1228 in a conserved acidic motif in the CD of Treacle (Mooser et al., 2020). However, we showed that Treacle with intact phosphoserines 1227/1228 but unable to undergo condensation loses the ability to interact with TOPBP1 during genotoxic stress. This observation implies that Treacle’s interaction with TOPBP1 and the subsequent activation of nucleolar DDR requires Treacle’s phase separation ability.

In conclusion, our study revealed Treacle’s role not only as a structural scaffold for the FC/DFC but also as a nucleolar hub protein that integrates functions in rDNA transcription, rRNA processing, and rDNA repair.

Materials and methods

Cell culture and drug treatment

Human HeLa (ATCC® CCL-2™), MCF7 (ATCC® HTB-22) and human skin fibroblasts (female 46XX) were cultured in DMEM (PanEco) supplemented with 10% fetal bovine serum (FBS; HyClone/GE Healthcare) and penicillin/streptomycin. The cells were cultured at 37°C in a conventional humidified CO2 incubator. DNA damage was induced by the treatment of cells with 90 µM etoposide (Sigma-Aldrich, #E1383) for 30 min or 1 h. For rRNA transcription inhibition, cells were treated with 0,05 µg/ml actinomycin D (Sigma-Aldrich, #A1410) or 20 µM CX-5461 for 3 h. To obtain 1,6-hexanediol (HD)-treated cells, HeLa cells were incubated with 5% 1,6-HD (Sigma-Aldrich, #240117) in serum-free medium at 37°C in a humidified atmosphere for 10 min. To obtain ammonium acetate-treated cells, HeLa cells were incubated with 200 mM ammonium acetate in a complete culture medium at room temperature for 5 min.

Plasmid constructs

For the FusionRED-Treacle-Cry2 (opto-Treacle) construct, the full-length Treacle was amplified by PCR from cDNA with primer set #1 (Supplementary Table 1) using KAPA High-Fidelity DNA Polymerase (KAPA Biosystems, KE2502). The forward and reverse primers contained XhoI sites. The amplified fragment was inserted into the FusionRed-C vector (Evrogen, FP411) linearized with XhoI. The Cry2 fragment was amplified by PCR from the plasmid pHR-mCh-Cry2WT (Addgene, #101221) with primer set #2 (Supplementary Table 1) digested with NheI and inserted into pFusionRed-Treacle linearized with NheI using NheI enzyme.

For the FusionRED-FUS-Cry2 construct, the FUS was amplified by PCR from the plasmid pHR-FUSN-mCh-Cry2WT (Addgene, #101223) with primer set #3 (Supplementary Table 1) using KAPA High-Fidelity DNA Polymerase (KAPA Biosystems, KE2502). The forward and reverse primers contained BspEI and KpnII sites respectively. The amplified fragment was inserted into the pFusionRed-C vector (Evrogen, FP411). The Cry2 fragment was amplified by PCR from the plasmid pHR-mCh-Cry2WT (Addgene, #101221) with primer set #2 (Supplementary Table 1) digested with NheI and inserted into pFusionRed-Treacle using NheI enzyme.

To generate Treacle-GFP or Treacle-Katushka2S constructs, the full-length Treacle was amplified by PCR from cDNA with primer set #4 (Supplementary Table 1) using KAPA High-Fidelity DNA Polymerase (KAPA Biosystems, KE2502). The forward and reverse primers contained BglII and BamHI sites respectively. The amplified fragment was inserted into the pTurboGFP-C (Evrogen, FP511) or pKatushka2S-C (Evrogen, FP761) vectors using BglII/BamHI restriction/ligation.

Treacle Δ1-83 deletion mutant was constructed based on the pTreacle-GFP full or pTreacle-2S full plasmid using iProof High-Fidelity DNA polymerase with primer set #5 (Supplementary Table 1). The resulting DNA template after PCR was reprecipitated and treated with the DpnI restriction enzyme. In the next step, the desired DNA template was purified on an agarose gel, phosphorylated with T4 Polynucleotide Kinase (T4 PNK), and ligated.

To obtain the charge-scrambled Treacle-GFP or Treacle-2S (Treacle-GFP CS or Treacle-2S CS), a fragment encoding 137-1130 aa of Treacle was removed from the pTreacle-GFP or pTreacle-2S plasmid using EcoR1 and HindIII restrictases. This fragment was then replaced with a synthetic sequence encoding the amino acid sequence with the necessary changes (see Supplementary Table 2).

To obtain the Treacle ΔSE mutant, a fragment encoding 137-1130 aa of Treacle was removed from the pTreacle-GFP or pTreacle-2S plasmid using EcoR1 and HindIII restrictases. This fragment was then replaced with a synthetic sequence encoding the amino acid sequence with the necessary changes (see Supplementary Table 2).

Treacle Δ83-1121 deletion mutant was constructed based on the pTreacle-GFP full or pTreacle-2S full plasmid using iProof High-Fidelity DNA polymerase with primer set #6 (Supplementary Table 1). The resulting DNA template after PCR was reprecipitated and treated with the DpnI restriction enzyme. In the next step, the desired DNA template was purified on an agarose gel, phosphorylated with T4 Polynucleotide Kinase (T4 PNK), and ligated.

Treacle Δ1121-1488-NLS mutant was constructed based on the pTreacle-GFP full or pTreacle-2S full plasmid using iProof High-Fidelity DNA polymerase with primer set #7 (Supplementary Table 1). The resulting DNA template after PCR was reprecipitated and treated with the DpnI restriction enzyme. In the next step, the desired DNA template was purified on an agarose gel, phosphorylated with T4 Polynucleotide Kinase (T4 PNK), and ligated.

For the generation of Treacle ΔNoLS (Δ1350-1488) mutant, fragments were amplified by PCR from the pTreacle-GFP plasmid with primer set #8 (Supplementary Table 1). The forward and reverse primers contained BglII and BamHI sites respectively. The amplified fragment was inserted into either the pTurboGFP-C (Evrogen, FP511) or pKatushka2S-C (Evrogen, FP761) vectors using BglII/BamHI restriction/ligation.

The full-length Treacle without fluorescent protein was constructed based on the pTreacle-2S full plasmid using iProof High-Fidelity DNA polymerase with primer set #9 (Supplementary Table 1). The resulting DNA template after PCR was reprecipitated and treated with the DpnI restriction enzyme. In the next step, the desired DNA template was purified on an agarose gel, phosphorylated with T4 Polynucleotide Kinase (T4 PNK), and ligated.

Fragment of the wt Treacle (amino acids 291–426) were amplified by PCR with primer set #10 (Supplementary Table 1) and subcloned into pMAL (New England Biolabs) vector encoding the TEV protease cleavage site after MBP. cDNA fragments and the expression vector were cleaved with BamHI I and Sal I digestion and ligated under suitable conditions.

The resulting all of constructs were verified by sequencing.

Gene knockdown

RNA interference experiments were performed using Lipofectamine 3000 transfection reagent (Thermo Scientific) following the manufacturer’s instructions. The cells were transfected with 50 nM Treacle/TCOF1 siRNA (the sequences of the siRNA is provided in Supplementary Table S3) or 50 nM TOPBP1 (Santa Cruz Biotechnology, #sc-41068).

For CRISPR/Cas9-mediated knockout, two single guide RNAs (sgRNA) to first exon of the TCOF1 gene were designed using the guide RNA design tool (www.atum.bio/eCommerce/cas9/input). The sgRNA targeting sequences were separately cloned into the px330mCherry (Addgene #98750). A list of all oligonucleotides is provided in Supplementary Table S3. The plasmids were co-transfected into HeLa cells with LTX transfection reagent (Invitrogen). 5-7 days after transfection cells were fixed and immunostained with required antibodies.

Fluorescence microscopy

For immunostaining, cells were grown on microscope slides. All samples were fixed in 1% formaldehyde in PBS for 15 min at room temperature and treated with 1% Triton X-100 for permeabilization. Cells were washed in PBS and then incubated with antibodies in PBS supplemented with 1% BSA and 0.05% Tween 20 for 1 h at room temperature or overnight at 4°C. Then the cells were washed with PBS three times (5 min each time). The primary antibodies bound to antigens were visualized using Alexa Fluor 488-conjugated secondary antibodies. The DNA was counterstained with the fluorescent dye 4,6-diamino-2-phenylindole (DAPI) for 10 min at room temperature. The samples were mounted using Dako fluorescent mounting medium (Life Technologies). The immunostained samples were analyzed using a Zeiss AxioScope A.1 fluorescence microscope (objectives: Zeiss N-Achroplan 40 ×/0.65 and EC Plan-Neofluar 100 ×/1.3 oil; camera: Zeiss AxioCam MRm; acquisition software: Zeiss AxioVision Rel. 4.8.2; Jena, Germany) or STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2). The images were processed using ImageJ software (version 1.44) and analyzed using CellProfiler software (version 3.1.5). 3D reconstruction of xyz confocal datasets (z-stacks) was performed using Leica LAS-X software. Antibodies used in the study are listed in Supplementary Table 7.

Cell sorting

Cells were transiently transfected with required plasmids using LTX transfection reagent (Invitrogen) or immunostained with required antibodies. Cell sorting was performed using an SH800 Cell Sorter (Sony) with a laser tuned to 488 nm for green fluorescence and 561 nm for red fluorescence. Gates were set with reference to negative controls. A minimum of 2×106 events was collected for ChIP or RNA extraction.

RNA extraction and RT-qPCR

After sorting, living or immunostained cells were pelleted by centrifugation at 3000 g for 5′ at 4°C. The supernatant was discarded. Total RNA was isolated from the pellet as described (Hrvatin et al, 2014). All RNA samples were further treated with DNase I (Thermo Scientific) to remove the residual DNA. RNA (1 µg) was reverse transcribed in a total volume of 20 µl for 1 h at 42°C using 0.4 µg of random hexamer primers and 200 U of reverse transcriptase (Thermo Scientific) in the presence of 20 U of ribonuclease inhibitor (Thermo Scientific). The cDNA obtained was analyzed by quantitative polymerase chain reaction (qPCR) using the CFX96 real-time PCR detection system (Bio-Rad Laboratories). The PCRs were performed in 20 µl reaction volumes that included 50 mM Tris–HCl (pH 8.6), 50 mM KCl, 1.5 mM MgCl2, 0.1% Tween-20, 0.5 µM of each primer, 0.2 mM of each dNTP, 0.6 µM EvaGreen (Biotium), 0.2 U of Hot Start Taq Polymerase (Sibenzyme) and 50 ng of cDNA. Primers used in the study are listed in Supplementary Table 4.

Chromatin immunoprecipitation (ChIP) and ChIP-seq analysis

Living cells were fixed for 15 min with 1% formaldehyde at room temperature, and crosslinking was quenched by adding 125 mM glycine for 5 min. Cell sorting was performed if needed. Cells were harvested in PBS, and nuclei were prepared by incubation in FL buffer (5 mM PIPES, pH 8.0, 85 mM KCl, 0.5% NP40) supplemented with Protease Inhibitor Cocktail (Bimake) and Phosphatase Inhibitor Cocktail (Bimake) for 30 min on ice. Next, chromatin was sonicated in RIPA buffer (10 mM Tris–HCl, pH 8.0, 140 mM NaCl, 1% Triton X-100, 0.1% sodium deoxycholate, 0.1% SDS) with a VirSonic 100 to an average length of 200-500 bp. Per ChIP reaction, 10–20 µg chromatin was incubated with 2–4 µg antibodies overnight at 4°C. The next day, Protein A/G Magnetic Beads (Thermo Scientific) were added to each sample and incubated for 4 h at 4°C. Immobilized complexes were washed two times for 10 min at 4°C in low salt buffer (20 mM Tris-HCl, pH 8.0, 150 mM NaCl, 2 mM EDTA, 0.1% SDS, 1% Triton X-100) and high salt buffer (20 mM Tris–HCl, pH 8.0, 500 mM NaCl, 2 mM EDTA, 0.1% SDS, 1% Triton X-100). Samples were incubated with RNase A (Thermo Scientific) for 30 min at room temperature. The DNA was eluted from the beads and de-crosslinked by proteinase K digestion for 4 h at 55°C and subsequent incubation at 65°C for 12 h. Next, DNA was purified using phenol/chloroform extraction and analyzed by qPCR. The qPCR primers used for ChIP analysis are listed in Supplementary Table 5. The sequencing libraries were then prepared with NEBNext Ultra II kit according manufacturer’s protocol. Final libraries were PCR amplificated and adapter dimers were cleaned with 1:1 MagPure magnetic beads (Magen Biotechnology). Resulted DNA was resuspended in 30 mkl 10 mM Tris-HCl buffer ph 8.0 and were sequenced on Illumina machine. Chip-seq reads were mapped to the reference human genome hg38 assembly using Bowtie v2.2.3 with the ‘–very-sensitive’ mode. Non-uniquely mapped reads, possible PCR and optical duplicates were filtered out using SAMtools v1.5. The bigWig files with the ratio of RPKM normalized ChIP-seq signal to the input were generated using deepTools v3.4.2 bamCompare function. Cool files was generated, merged and normalized using cooler version 0.8.11 (https://github.com/open2c/cooler).

Electron microscopy

Twenty-four hours after transfection, cells were fixed with 2.5% neutralized glutaraldehyde in the requisite buffer for 2 h at room temperature, post-fixed with 1% aqueous OsO4, and embedded in Epon. Sections of 100 nm thickness were cut and counterstained with uranyl acetate and lead citrate. Sections were examined and photographed with a JEM 1400 transmission electron microscope (JEOL, Japan) equipped with a QUEMESA bottom-mounted CCD-camera (Olympus SIS, Japan) and operated at 100 kV.

Protein purification

E. coli BL21 (DE3) cells were transformed with vectors encoding cDNAs of TCOF LCR fragments fused with TEV-cleavable maltose-binding protein (MBP). BL21 (DE3) were cultured in Terrific Broth medium at 37°C to OD=0.6-1.0. Protein expression was induced by adding 0.8 mM IPTG, and the cultures were further incubated at 18 °C overnight. The cells were collected by centrifugation (5,000g, 20 min, 4 °C). To purify MBP-tagged protein under a native condition, the cell pellet was dissolved in start buffer (20 mM Tris–HCl, 150 mM NaCl, 10 mM MgCl2, 0.1% NP-40, 10% glycerol, 1 mM DTT, 1 mM phenylmethylsulfonyl fluoride, 1x protease inhibitor cocktail (B14001 selleckchem), pH 7.4). Cells were disrupted by sonication in start buffer on ice. The lysate was centrifuged (20,000g, 4 °C, 60 min) and the supernatant was collected and applied to a column containing amylose resin (NEB, E8021S) at 4 °C. The column was washed with wash buffer (20 mM Tris–HCl, 500 mM NaCl, 10 mM MgCl2, 0.1% NP-40, 10% glycerol, 1 mM DTT, pH 7.4).

For cleavage of MBP TEV protease was added at molar ratio approximately 1:50 directly to the eluted protein in buffer (20 mM Tris–HCl, 200 mM NaCl, 5 mM sodium citrate, 1 mM DTT, pH 7.4) overnight at 4 °C while gently mixing the resin on a rotator. Subsequently, the tagged proteins were eluted and then subjected to ion-exchange chromatography with 0-1000 mM NaCl gradient.

Eluted proteins were sequentially dialysed at 4 °C against assay buffer (20 mM Tris–HCl pH 7.4) at 4 °C overnight and concentrated using Amicon centrifugal filters (Millipore) and stored at −80 °C in small aliquots.

In vitro Treacle condensation analysis

Recombinant Treacle fragment (amino acids 291–426) was diluted in 20 mM Tris–HCl pH 7.4 to a final concentration of 100 µM. To assemble condensates PEG 8000 was added to the protein solution to a final concentration of 5%, followed by incubation at room temperature for 10 minutes. 10 μL of the resulting solution was pipetted onto a glass slide, covered with a coverslip, and analyzed on the Zeiss AxioScope A.1 fluorescence microscope in DIC mode. To test the effects of NaCL or 1,6-hexandiol on Treacle condensation, NaCL or 1,6-hexandiol was added to the protein solution with 5% PEG 8000 at 500 mM or 10% concentrations, respectively.

Live-cell imaging

Cells were seeded in 35-mm glass-bottom dishes twelve hours before transfection with the plasmid. Twenty-four hours after transfection, Hoechst 33342 (Cell Signaling Technology) was added to the medium at a final concentration of 1 μg/ml, and the cells were incubated for 20 min at 37°C. Hoechst-containing medium was replaced with a complete fresh medium before the cells underwent live-cell imaging using a STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2) equipped with an incubation chamber to provide a humidified atmosphere at 37°C with 5% CO2. For observation of liquid droplet behavior, z-stack time-lapses were taken. Maximum intensity projections are shown, but all findings were confirmed on a single plane.

Optodroplet assay

Cells were seeded in 35-mm glass-bottom dishes and were transfected with Treacle-FusionRed-Cry2 plasmid using LTX transfection reagent (Invitrogen) 24 h before imaging. Twenty-four hours after transfection, Hoechst 33342 was added to the medium at a final concentration of 1 μg/ml, and the cells were incubated for 10 min at 37°C. Hoechst-containing medium was replaced with a complete fresh medium before the cells underwent live-cell imaging using a STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2) equipped with an incubation chamber to provide a humidified atmosphere at 37°C with 5% CO2. Droplet formation was induced with light pulses at 488 nm (blue light, 1% laser power) for 10 sec, and z-stack images were captured every 60 sec in the absence of blue light.

Fluorescence recovery after photobleaching (FRAP) analysis

Fluorescence recovery after photobleaching (FRAP) experiments were performed on a STELLARIS 5 Leica confocal microscope equipped with an HC PL APO 63x/1.40 oil CS2 objective and an incubation chamber that maintained a humidified atmosphere at 37°C with 5% CO₂. Live-cell experiments were performed using transfected cells cultured in phenol red-free DMEM medium supplemented with 10% fetal bovine serum. Imaging was typically performed at a resolution of 512 × 512 pixels, with a scan speed of 200 ms per frame.

Typically, images were acquired at 512 × 512 pixels at a scan speed corresponding to 200 ms per image. Before photobleaching, 3 to 5 images were recorded. Bleaching parameters, i.e., laser intensity and scanning time, were chosen to reach approximate 50% of bleaching in the shortest possible time. The bleaching area was selected according to the type of experiment (full-, partial- or half-FRAP). In the case of half-FRAP, which relies on the analysis of the bleached and the non-bleached half, it is important to optimize the bleaching step (intensity and area of the bleaching spot) to minimize bleaching in the other (non-bleached) half. Half-FRAP experiments were conducted until the signals in both halves had converged to each other or until the signal in the non-bleached half had reached its initial pre-bleach value. FRAP data analysis was performed as described in (Muzzopappa et al., 2022).

Proximity ligation assay

After drug treatments or at 24 h post-transfection, cells were fixed for 15 min with 1% formaldehyde at room temperature and permeabilized with 1% Triton X-100. Next, cells were subjected to the proximity ligation assay (PLA) using the Duolink Green (Sigma-Aldrich) according to the manufacturer’s instructions. Briefly, cells were blocked, incubated with appropriate primary antibodies overnight at 4°C, and then incubated with PLA probes, which are secondary antibodies (anti-mouse-IgG and anti-rabbit-IgG) conjugated to unique oligonucleotides. Samples were then treated with a ligation solution followed by an amplification solution containing polymerase and fluorescently labeled oligonucleotides, allowing rolling-circle amplification and detection of discrete fluorescent dots. Some samples after the PLA protocol were additionally counterstained with Alexa-Fluor-488- or Alexa-Fluor-594-conjugated (Molecular Probes) secondary antibody. The DNA was counterstained with DAPI for 10 min at room temperature. The samples were mounted using Dako fluorescent mounting medium (Life Technologies) and analyzed using STELLARIS 5 Leica confocal microscope (objectives: HC PL APO 63x/1.40 oil CS2). The images were analyzed using ImageJ software (version 1.44).

Transcription labeling

For EU incorporation, the cells were incubated with 100 μM EU (Sigma-Aldrich) for 2 hr at 37°C. Then, the cells were washed three times with PBS and fixed in ice-cold methanol for 10 min. The samples were then processed using a Click-iT EU Imaging Kit (Life Technologies) according to the manufacturer’s recommendations.

Single molecule RNA Fluorescentin in situ Hybridization (smFISH)

All single-molecule RNA FISH probes were designed as described (Yao et al., 2019) and labeled with FITC on the 3’ends (Supplementary Table 6). Cells were fixed with 4% PFA for 15 min, followed by permeabilization with 1% Triton X-100 for 10 min. Cells were incubated in 10% formamide/2x SSC for 10 min at room temperature and then were hybridizated with 5 nM each of RNA probes in 50% formamide/2x SSC at 37◦C for 16 hours. After hybridization, the cells were washed 10% formamide/2x SSC for 30 min at 37◦C.